Chromosome-scale genome assembly of the European common cuttlefish Sepia officinalis

Version 1

Abstract

Abstract

Coleoid cephalopods, a subclass of mollusks, exhibit remarkable adaptations, including the largest brains among invertebrates, camera-like eyes, and a distinctive embryonic development. They possess an advanced behavioral repertoire including dynamic camouflage.

The common cuttlefish Sepia officinalis has served as a key model organism in various research fields, spanning biophysics, neurobiology, behavior, evolution, ecology and biomechanics. More recently, it has become a model to investigate the neural mechanisms underlying cephalopod camouflage, using quantitative behavioral approaches alongside molecular techniques to characterize the identity, evolution and development of neuronal cell types.

Despite significant interest in this animal, a high-quality, annotated genome of its species is still lacking. To address this, we sequenced and assembled a chromosome-scale genome for S. officinalis. The final assembly spans 5.68 billion base pairs and comprises 47 repeat-rich chromosomes. Gene linkage analysis confirms the existence of 47 chromosomes, revealing clear homologies with related species such as Euprymna scolopes and Doryteuthis pealeii. Our work includes a comprehensive gene annotation and full-length transcript predictions that should be helpful for further evolutionary and single-cell expression studies. This genome provides a valuable resource for future research on the evolution, brain organization, information processing, development, and behavior in this important clade.

Introduction

Coleoid cephalopods (octopus, squid, cuttlefish) are a highly derived group of mollusks, characterized by the largest nervous systems among all invertebrates (ca. 500 million neurons in an adult octopus1,2, ca. 200 million neurons in adult cuttlefish) and specializations with a great historical importance for neuroscience (e.g., “giant axons”3 and “giant synapses”4–6). These animals exhibit very sophisticated behaviors such as dynamic camouflage7–13, learning14–16, social communication17–19 and hunting20–22 as well as two-stage sleep23–25. Because of these characteristics, cephalopods have been the focus of many fundamental studies by biologists, biophysicists and physiologists over the past century26–35. Thanks to advances in sequencing technologies, coleoid cephalopods are now also emerging as animal models in molecular neurobiology, evolution and genomics36. Recent studies in particular examined cephalopod biology from the perspectives of single-cell gene expression37–40, genome topology and gene regulation41.

Despite the technical progress, the genomes of cephalopods remain challenging to assemble because of their large sizes and high repeat fractions, creating a bottleneck for further molecular studies. While the genomes of Octopodiformes (Octopus, Eledone, Argonauta) are often comparable in size to that of humans (around 3 Gigabases (Gb)42,43) or possibly even smaller44, the typical genomes of Decapodiformes (squids and cuttlefish) are often twice as large, reaching 6 Gb45,46. The biggest contributing factors to this genome expansion are transposable elements (TEs), with different TE classes differentiating squid and cuttlefish genomes from those of Octopodiformes47. Besides such differences in genome sizes, karyotypes in coleoid cephalopods are also unusual when compared to those of other invertebrates, including mollusks: for example, haploid chromosome numbers are around 30 in octopuses and around 46 in squids and cuttlefish46,48, while they are between 8 and 19 in non-cephalopod mollusks49–56. Karyotype reconstruction and validation techniques in animals with such large genomes are also not as developed as they are for more familiar invertebrate and vertebrate species.

The common cuttlefish Sepia officinalis has been used as a model organism for studies in neural control and development57,58, behavior7,15, evolution59, environmental studies60,61 and biomechanics62,63. Found in the Mediterranean, the English Channel and the European Atlantic, it can produce large egg clutches containing hundreds of embryos, several times throughout its semelparous reproductive cycle64,65. More recently, this species has emerged as an important model to study the neural basis of cephalopod camouflage66,67 building on extensive behavioral studies7–13 and neuronal cell type characterization, evolution and development68–70.

Despite a widespread interest in this animal, including the recent sequencing of several cephalopod species as part of the Aquatic Symbiosis Genomics project71, a high-quality, annotated genomic resource for this species is still lacking. We describe here the sequencing, assembly, and annotation of the Sepia officinalis genome. Our analysis provides evidence for the existence of 47 chromosomes, with clear homologies to the chromosomes of related species such as Euprymna scolopes and Doryteuthis pealeii. We predict full-length transcripts and provide functional gene annotation that should inform future evolutionary and transcriptomic studies. Taken together, this genome should be an important resource for scientists interested in genome evolution, brain organization, information processing, development and behavior.

Results

We sequenced the genome of a 6-month-old Sepia officinalis individual (probably male, see below), reared from eggs collected in the Portuguese Atlantic. The DNA was extracted from brain tissue and sequenced using long-read (PacBio HiFi) and chromatin conformation (Hi-C) methods (Figure 1A,B).

Sepia officinalis assembly statistics and quality control.

A) Specimen of S. officinalis (credit: Stephan Junek, MPI Brain Research). B) Overview of the assembly workflow. Genome size was estimated using GenomeScope72,73. The primary assembly was generated from long DNA reads (PacBio Sequel II) and HiC reads (Dovetail OmniC) with hifiasm124. Assembly was scaffolded with YAHS77. C) Snail plot of chromosome-scale S. officinalis assembly generated using blobtools2135 showing scaffold statistics, base composition and BUSCO75 completeness against the metazoa_odb10 database. D) HiC heatmap showing 47 chromosome-scale scaffolds with very few sequences in unplaced scaffolds. Heatmap was generated using juicebox78, 0-7039 observed counts (balanced) are shown.

Genome size and heterozygosity

The haploid genome size of S. officinalis was estimated to be ∼5.14 Gb based on k-mer estimation from the long-read data, with a high repeat content of 54 %72,73. The heterozygosity rate was estimated to be 1.03 %, which is higher than in octopus genomes, yet moderate among marine invertebrates42,43,46.

The size of the scaffolded assembly was 5.68 Gb (Figure 1C), thus about 10% greater than our initial GenomeScope estimate. Indeed, the sizes of most published metazoan genome assemblies deviate by more than 10% from estimates, with proportionally greater deviations as assembly sizes increase74.

Completeness of the genome was assessed using BUSCO75 with metazoa_odb10. It was 95.2% [single copy: 93.5%, duplicated: 1.7%], with 3.2% fragmented and 1.6% missing BUSCOs. The BUSCO score with mollusca_odb10 for the final assembly was 77.2% completeness [single copy: 75.9%, duplicated: 1.3%] with 4.3% fragmented and 18.5% missing BUSCOs.

Karyotype

The assembly was generated from PacBio HiFi reads and Hi-C reads with ca. 23-fold and 11-fold coverage, respectively. Scaffolding was performed using Hi-C reads, placing 95.8 % of bases in 47 scaffolds. The Hi-C heatmap in Figure 1D shows the chromatin contacts in clusters corresponding to these 47 scaffolds. It is consistent with the proposal that they each represent a single chromosome.

To increase confidence in our assembled Sepia officinalis karyotype, we used several scaffolding programs and manually curated the scaffolds using two independent approaches. The first used HapHiC76, a tool based on Hi-C data that utilizes allele information from primary assembly programs, and scaffolds assemblies with the constraint of an expected number of chromosomes. The resulting Hi-C contact maps with 46, 47, 48, 49 and 50 chromosomes (Supplementary Figure 1) show support for 47 chromosomes; an input of 46 scaffolds resulted in one clear chromosome merger (blue arrowhead); higher input numbers resulted in false chromosome splits (black arrowheads, 48, 49 and 50 scaffolds).

Second, we performed manual curation of the scaffolding result obtained with YAHS77 using JBAT78. These two approaches, conducted independently and both based on Hi-C signal, converged on 47 chromosomal scaffolds.

Karyotype comparisons with other Decapods

Previous karyotyping studies in decapod cephalopods based on cytogenetic techniques provide widely varying estimates of chromosome numbers, reflecting the difficulty of resolving large chromosome numbers in situ. For example, 1n=46 chromosomes has been reported for two species of cuttlefish (Sepia esculenta and Sepia lycidas) and three loliginid squids79; 1n=34 chromosomes has been reported for Sepia arabica80 and 1n=24 chromosomes in Sepia pharaonis81. In Sepia officinalis, a karyotype of 1n=52 has even been described from testis samples82. More recently, several cephalopod genomes have been scaffolded at chromosome-scale resolution, resolving karyotypes in silico. From those recent studies, a haploid karyotype of 46 chromosomes seems to be common among Decapodiformes46 (Figure 2A).

Syntenic comparison of three decapod species.

A) Taxonomy of selected cephalopod species showing their genome size and (haploid) chromosome numbers. Taxonomy information downloaded from NCBI taxonomy browser, divergence times from85,136. B) Synteny relationship between chromosomes of E. scolopes46 (top), D. pealeii46 (middle) and S. officinalis (bottom). Riparian plot was generated using orthogroups located on chromosomes of the species. Euprymna chromosomes 45 and 46 are not shown because they contain too few orthogroups. C) Synteny plot of Sepia chromosomes 40 (magenta) and 43 (darkblue), that are joined in the other species and cause the different haploid chromosome number in Sepia.

Comparison with another chromosome-scale Sepia officinalis assembly

A chromosome-scale assembly for Sepia officinalis was released recently by the Wellcome Sanger Institute’s Darwin Tree of Life project83 (DToL, GCA_964300435.1). That genome was assembled from a male individual using PacBio Sequel II and Arima2 Hi-C data, with a final assembly size of 5.8 Gb. In that study, the haploid chromosome number was estimated to be 49. We scaffolded their assembly using our Dovetail Omni-C data, and the resulting contact map again supports 47 chromosomes (Supplementary Figure 2A), consistent with our results. We next aligned the two S. officinalis assemblies using minimap284 to assess the differences between them (Supplementary Figure 2B). The “Sanger” assembly appears to split three chromosomes of our assembly (chromosomes 1, 5 and 42) and to merge two (chromosomes 40 + 43). In conclusion, available evidence supports a haploid karyotype of 47 chromosomes in Sepia officinalis.

Comparisons with other chromosome-scale Decapodiformes genomes

To further investigate the possible reasons why our estimate of Sepia officinalis’s chromosome numbers deviates from those in other studies, we compared our assembly with the chromosome-scale genomes of two other decapod cephalopods, Euprymna scolopes41 (a sepiolidae) and Doryteuthis pealeii46 (a loliginidae), both described as having 46 chromosomes. Using orthogroups to perform linkage analysis, we could detect a clear chromosome homology between Doryteuthis and Sepia, and to a lesser extent between Euprymna and Sepia (Figure 2B). Dotplots showing detailed pairwise synteny comparisons are shown in Supplementary Figure 3 (Sepia to Doryteuthis) and Supplementary Figure 4 (Sepia to Euprymna). The inferred species tree as part of this analysis places Euprymna as sister to Sepia and Doryteuthis, matching recent phylogenetic analyses50,85. However, this tree was constructed without the inclusion of outgroup taxa, and therefore lacks a reliable root.

This comparison revealed that chromosomes 40 and 43 in Sepia are merged into one in Euprymna and in Dorytheutis (Figure 2C), as they were in the DToL Sepia officinalis assembly (Supplementary Figure 2B). Our scaffolding approach showed no contact between chromosomes 40 and 43 in our Sepia Hi-C data, providing support for these chromosomes being separate in Sepia officinalis.

We also compared our assembly to a chromosome-scale Sepia esculenta assembly (1n=46, GCA_964036315.1), and recently annotated by genome liftover from Sepia pharaonis86. That assembly, constructed from a female individual, showed a low read coverage for a putative sex chromosome in a ZZ/Z0 sex determination system87. Our syntenic comparison indicated a strong homology between the putative Z chromosome of Sepia esculenta and chromosome 46 of Sepia officinalis (Supplementary Figure 5A). Because we did not record the sex of the Sepia officinalis animal used for genome sequencing and assembly, we tried to infer the sex by replicating the analysis used to identify the putative Z chromosome in Sepia esculenta87. For this, we aligned short-reads (Illumina) from the same Sepia officinalis individual to the assembly and examined the normalized read coverage for each chromosome (Supplementary Figure 5B). In contrast to the low coverage observed in the female Sepia esculenta assembly (Supplementary Figure 5C), we observed no decrease in read coverage for chromosome 46, suggesting that, if the alleged sex chromosome identity is correct, then our material came from a male animal.

Genome repeat landscape

After creating a custom repeat library for Sepia officinalis using RepeatModeler88, we masked the genome using RepeatMasker89, resulting in 71.17% masked bases. The categories of repeats are shown in Figure 3A. Most repeats (39.65% of total bp) were not characterized and presumably represent ancient repeats that diverged beyond recognition90. Retroelements constituted the largest characterized repeat category (17.32%) followed by DNA transposons (5.92%); 5.83% were annotated as simple repeats. As observed in other Decapodiformes, the Sepia genome contained almost no short interspersed nuclear elements (SINEs), supporting the hypothesis that the SINE expansion observed in octopuses47 occurred independently in their lineage.

Genome annotation for Sepia officinalis.

A) Annotation of repeat landscape of the S. officinalis genome, annotated using RepeatModeler88. Full repeat landscape is shown on the left, annotated repeats (excluding unclassified or simple repeats) are shown on the right. B-C) Quality control of gene annotation and comparison to two other cuttlefish species using OMArk109. Results shown for Sepia lycidas (GCA_963932145.1, Ensembl Genebuild), Sepia officinalis (BRAKER, this study) and Sepia pharaonis137 (BRAKER). Lophotrochozoa was used as the ancestral clade. B) Completeness assessed by the presence of genes conserved in the clade, classified as single or multiple copies (duplicated), or missing. C) Consistency assessed by the proportion of proteins placed in the correct lineage (consistent); placement in incorrect lineages randomly (inconsistent) or to specific species (contamination), or no placement in known gene families (unknown).

Gene modeling and annotation

The genome was further annotated using BRAKER75,91–106 combining short- and long-read RNA-seq data and publicly available protein data from multiple molluscan species (Doryteuthis pealeii46, Euprymna scolopes107, Octopus bimaculoides46, Octopus vulgaris43, Nautilus pompilius108, and Pecten maximus51). A total of 18,663 gene models and 23,768 proteins were annotated. The gene annotation was evaluated using BUSCO75 v5.5.0 in protein mode, showing high completeness of the annotation (metazoa_odb10: C:98.2%[S:77.3%,D:20.9%], F:1.0%, M:0.8%, n:954 and mollusca_odb10: C:81.5%[S:59.6%,D:21.9%], F:0.9%, M:17.6%, n:5295). We further checked the completeness and consistency of our gene models using OMArk109, and compared them to genome annotations for two other cuttlefish species, Sepia pharaonis and Sepia lycidas (Figure 3B). The completeness is assessed by comparing the annotated genes to conserved orthologs (present in 80% of extant species) in a given taxonomic clade, here the superorder Lophotrochozoa. The S. officinalis annotation misses 181 out of 2373 genes, which is higher than S. lycidas (96 genes) but much lower than S. pharaonis (458 genes).

In a consistency assessment, where the presence of known gene families from the lineage is evaluated (Figure 3C), S. officinalis contains low proportions of taxonomically inconsistent or fragmented proteins (ca. 13%) similar to S. lycidas (9%), giving confidence in the annotation. In contrast, more than 50% of S. pharaonis proteins are labeled as inconsistent or fragmented, which could indicate annotation errors109.

Overall, the annotations contain different numbers of predicted proteins, likely caused by differences in the annotation method and reference data used. S. officinalis contains the least proteins (23,768) compared to S. lycidas (35,949) and S. pharaonis (53,515). The S. officinalis annotation thus represents a conservative yet high-quality gene set. In the future, this annotation could be enhanced by incorporating more long-read RNA sequencing data, for example.

Lastly, we assigned orthology information to the S. officinalis proteome from InterPro110 and eggNOG-mapper111 to aid the interpretability of the resource for future transcriptomic or proteomic studies. Overall, 89% of proteins (21,204 out of 23,768) received an annotation from InterPro from at least one of their databases. 59% of proteins (14,126 out of 23,768) were annotated by eggNOG-mapper, reflecting the more stringent orthology filters and prioritization of full-length matches implemented in the program112. Similar to previous studies of other cephalopod genomes42, we observed a high number of C2H2 zinc finger transcription factors (1,538 proteins, InterPro annotation) and protocadherins (339 proteins, InterPro), suggesting a conserved expansion of these gene families among cephalopods. Further analyses should reveal more expanded gene families in S. officinalis that underlie species-specific adaptations.

Discussion

This study presents a high-quality, chromosome-scale genome assembly and annotation for the common cuttlefish Sepia officinalis, providing an additional important resource for comparative genomics, neurobiology, and evolutionary studies within cephalopods and mollusks. The assembly was derived from a combination of PacBio HiFi long reads, Dovetail Omni-C chromatin conformation capture, and multiple rounds of scaffolding and manual curation, resulting in a genome size of 5.68 Gb and a haploid karyotype of 47 chromosomes. This resource represents a significant step forward for cephalopod genomics, especially within the Decapodiformes, where large and repetitive genomes have hindered previous efforts.

Chromosome Number and Karyotypic Variation

Our results strongly support a haploid chromosome number of 47 in S. officinalis, a conclusion reinforced by two independent scaffolding approaches (YAHS77 and HapHiC76) and supported by Hi-C contact maps. This karyotype contrasts with the 46 chromosomes reported in other decapod cephalopods, such as Euprymna scolopes and Doryteuthis pealeii46. Syntenic comparisons revealed that chromosomes 40 and 43 in S. officinalis correspond to a single chromosome in these other species. The current Hi-C data for Sepia officinalis provides clear evidence that chromosomes 40 and 43 are distinct, with no interaction contact. Because these two chromosomes are merged also in S. esculenta, it is plausible that this difference results from a S. officinalis-specific fission. However, our results show how difficult it still is to assemble large genomes with high karyotype numbers. Therefore, validation and comparisons with other Sepia species’ genomes will be important. This, in turn, will help identify chromosomal fusion-with-mixing events that are clade specific. Early branching order in Decapodiformes has been notoriously unstable50,85,113–116; thus, such rare and irreversible phylogenetic characters could be useful in further phylogenetic analysis of this clade48,117.

We also compared our results to the recently published S. officinalis genome from the Darwin Tree of Life project (DToL)83, which proposed 49 chromosomes. The discrepancy is attributable to one merger in our assembly and three apparent chromosome splits in the DToL assembly. While these differences could stem from technical choices such as different Hi-C methods (Arima v2 vs. Dovetail Omni-C) and sequencing depth (83-fold vs. 11-fold coverage), our assembly is strongly supported by Hi-C contact data, especially in regions where the DToL assembly splits chromosomes that appear contiguous in our data. Interestingly, the DToL assembly merges the chromosomes 40 and 43 - supporting a 46-chromosome model as ancestral state. However, this is not supported by re-scaffolding with the current Omni-C dataset, which again shows them as separate. This discrepancy might reflect technical differences, but it could also be hinting at population-level variability in chromosomal architecture; that is, some individuals may carry a fusion while others do not. Intraspecific chromosome number variation is rare but not unprecedented; for instance, chromosomal polymorphism has been described in the butterfly Leptidea sinapis across different populations118,119. Notably, the two S. officinalis individuals used for sequencing originated from different regions - the Portuguese Atlantic (this study) and the French Mediterranean (DToL project), raising the possibility of geographic variation in chromosome structure. Future population-level analyses will be critical to determine whether S. officinalis exhibits karyotypic polymorphism.

Genome Size and Repeat Landscape

The size of the genome of S. officinalis is estimated at 5.68 Gb - comparable to other Decapodiformes and roughly double the size of typical Octopodiformes genomes. Repeat annotation revealed that over 71% of the genome is repetitive, dominated by unclassified elements and retrotransposons. The near absence of SINEs reinforces the idea that their expansion in octopuses was a lineage-specific event47. These results underscore the complexity and evolutionary dynamism of the cephalopod genome, shaped by waves of transposon activity that likely played a role in the evolution of novel traits41,46.

Gene Annotation and Comparative Assessment

We annotated 18,663 gene models and 23,768 proteins using a combination of short- and long-read RNA-seq data and molluscan protein references. Compared with other annotated cuttlefish genomes (S. lycidas, S. pharaonis), our annotation is conservative in gene number but stands out in terms of completeness and taxonomic consistency. OMArk-based evaluations confirmed that our gene models have low levels of fragmentation and contamination, and high lineage-specific coherence. Striking differences across species in the number of predicted genes, duplications, and missing orthologs likely reflect variation in the annotation pipelines and reference data rather than true biological divergence. Still, our dataset provides a high-confidence foundation that can be refined with additional transcriptomic data from long-read techniques and diverse tissues.

As in other coleoid genomes, we observed notable expansions of gene families such as C2H2 zinc finger transcription factors and protocadherins, genes implicated in neural development and plasticity42,45. These conserved expansions across cephalopods point to genomic underpinnings of their cognitive complexity and behavioral sophistication.

Conclusions and Outlook

This study delivers a foundational genomic resource for Sepia officinalis, a model for studying neural control, behavior, and adaptive camouflage. The chromosome-scale assembly and annotation open the door to in-depth studies of gene regulation, neural circuit development, and genome evolution across coleoid cephalopods. Moreover, the identification of possible population-level karyotype variation highlights the importance of incorporating broader sampling strategies in future genomic studies.

As we move toward a more complete picture of cephalopod genome evolution, the integration of chromosomal synteny, regulatory architecture, and transcriptomic diversity across species will be especially important. The S. officinalis genome represents an important step on this path, enabling high-resolution comparative and functional analyses of one of the most enigmatic and evolutionarily successful invertebrate lineages.

Materials & Methods

Animal husbandry

All research and animal care procedures were carried out following the institutional guidelines that comply with national and international laws and policies (DIRECTIVE 2010/63/EU; German animal welfare act; FELASA guidelines). European cuttlefish Sepia officinalis were hatched from eggs collected in the Portuguese Atlantic and reared in a seawater system, at 20°C. The closed system contains 4,000l of artificial seawater (ASW; Instant Ocean) with a salinity of 33‰ and pH of 8–8.5. Water quality was tested daily and adjusted as required. Trace elements and amino acids were supplied weekly. Marine LED lights above each tank provided a 12/12-h light/dark cycle with gradual on- and off-sets at 07:00 and 19:00. The animals were fed live food (either Hemimysis spp. or small Palaemonetes spp.) ad libitum twice per day. The animals were housed together in 120-l glass tanks with a constant water through-flow resulting in five complete water exchanges per hour. Enrichment consisted of natural fine-grained sand substrate, seaweed (Caulerpa prolifera), rocks of different sizes and various natural and man-made three-dimensional objects.

Tissue preparation

Animals underwent terminal anesthesia to prevent animal suffering or distress, following the Guidelines for the Care and Welfare of Cephalopods in Research120,121. Animals were transferred to a bucket containing ethanol 2% (v/v) in ASW. After 5 minutes, animals were gently probed for simple avoidance reflexes, and the ethanol concentration in the ASW was gradually increased to a maximum of 5%. Sufficient depth of anesthesia was reached when the animal no longer reacted even to stronger stimuli (touching and pinching with tweezers), and no reaction was observed when a hand was moved in the visual area and the cornea was touched122. In this deep anesthesia, the animals were decapitated and the tissue was rapidly transferred into ice-cold, calcium-free Ringer solution (460 mM NaCl, 10 mM KCl, 51 mM MgCl2, 10 mM glucose, 2 mM Glutamine, 10 mM HEPES, pH 7.4) bubbled with oxygen. Brain and body tissue was dissected under a stereoscope and flash-frozen in liquid nitrogen before storage at -80°C until further use.

Unless stated otherwise, sequencing libraries were prepared from the same individual (6-month-old adult Sepia officinalis, F1 from eggs collected in Portugal).

Long-read Whole Genome Library Preparation and Sequencing

The long-read library was prepared and sequenced at the MPI for Plant Research in Cologne, Germany. Genomic DNA was extracted from flash-frozen brain tissue with the MagAttract HMW DNA kit (Qiagen, Cat. no. 67563) and the sequencing library was prepared with the SMRTbell express template prep kit 2.0 (PacBio, Cat. no. 100-938-900). The library was sequenced on 5 SMRT cells on the PacBio Sequel II with the Sequel II binding kit 2.2 (102-089-000), Sequel II sequencing kit 2.0 (101-820-200) and SMRT Cell 8M tray (101-389-001).

Long-read RNA library preparation and sequencing

RNA was isolated from 5-10 mg of flash-frozen tissue (brain, mantle/epidermis, arm/tentacle). First, tissue samples were homogenized in TRIzol using a homogenizer (VDI12/S12N-5S, VWR, Germany). RNA was isolated and DNAseI treated according to the Direct-zol RNA Mini prep kit from ZymoResearch (R2050). RNA was quantified by spectrophotometry (Nanodrop, Thermo Scientific, USA) and its quality was assessed with a RNA 6000 Nanochip on Bioanalyzer (Agilent Technologies, Germany).

The Iso-Seq libraries were prepared and sequenced at the Sequencing facility of the MPI for Plant Research, using a method targeting the 5’ cap and poly-A tail of mRNAs, described in detail in123. Briefly, mRNA was pooled from the 8 tissue samples and cDNA was synthesized using the TeloPrime Full-Length cDNA Amplification Kit V2 (Lexogen, Cat Nr: 013.08 or 013.24). The sequencing library was prepared with the SMRTbell express template prep kit 2.0 (PacBio, Cat. no. 100-938-900). The pooled library was sequenced on 1 SMRT cell on the PacBio Sequel II system using the Sequel II binding kit 2.1 (101-843-000), Sequel II sequencing kit 2.0 (101-820-200) and SMRT Cell 8M tray (101-389-001).

Further, a separate library was prepared from optic lobe cDNA and sequenced on a second SMRT cell using the same reagents described above.

Omni-C Library Preparation and Sequencing

An Omni-C library was prepared and sequenced at the MPI for Plant Research in Cologne, Germany. The library was prepared from brain tissue using the DovetailTM Omni-C™ Kit (Dovetail Genomics, Cat. No. 21005) according to the manufacturer’s protocol. Briefly, chromatin was fixed in place in the nucleus. Fixed chromatin was digested with DNase I then extracted. Chromatin ends were repaired and ligated to a biotinylated bridge adapter followed by proximity ligation of adapter-containing ends. After proximity ligation, crosslinks were reversed and the DNA was purified from proteins. Purified DNA was treated to remove biotin that was not internal to ligated fragments. The sequencing library was generated using Illumina-compatible adapters. Biotin-containing fragments were isolated using streptavidin beads before PCR enrichment of the library. The library was sequenced on an Illumina NextSeq 2000 platform to generate 400 million 2 x 150 bp read pairs.

Short-read DNA sequencing

High molecular weight genomic DNA was isolated from brain tissue (optic lobe) using NEB Monarch gDNA Purification Kit (T3010S). First, 10 mg of flash-frozen tissue were submerged in 100 µl of tissue lysis buffer, cut into small pieces and transferred into a 1.5 ml tube containing an additional 100 µl of lysis buffer. Then, 3 µl of 20 U/µl Proteinase K was added to the tissue sample and incubated for 2 h at 56°C in a thermal mixer with agitation until tissue pieces were completely dissolved. After incubation, tissue lysate was purified from the debris by centrifugation for 3 min at maximal speed (Eppendorf, Germany). Clean supernatant was transferred to a new 1.5 ml tube and incubated with 3 µl RNase A for 5 min at 56°C with agitation at full speed. After RNase A incubation tissue lysate was column purified and eluted with 80 µl of elution buffer according to the manufacturer protocol. The purity of gDNA was assessed using A260/280 and A260/230 absorbance ratios measured by NanoDrop spectrophotometer (ThermoFisher Scientific). The integrity of the obtained gDNA was checked using 0,75% agarose gel electrophoresis by loading 100 ng of gDNA sample and using Hind III digested Lambda DNA (NEB#3012) as a marker.

DNA sequencing libraries were prepared from 500 ng of high MW gDNA using Illumina DNA PCR-Free Tagmentation Library Prep Kit (20041794) and UD Indexes Set A (20026121). Obtained dual-indexed single-stranded libraries were quantified using the Qubit single-stranded DNA (ssDNA) assay kit (Q10212, ThermoFisher Scientific). 1000 pM of the final library were run on P3 reagents and Illumina NextSeq2000 using Illumina DNA PCR-Free Sequencing and Indexing primer (20041797) with 151 bp paired-end reads.

Short-read RNA library preparation and sequencing

For short-read RNA sequencing, tissue from anoth er animal (8-month-old adult, F0 from eggs collected in Normandie, France) was used. RNA was isolated from 5 mg of flash-frozen tissues (brain, skin and retina). First, tissue samples were homogenized in TRIzol using a homogenizer (VDI12/S12N-5S, VWR, Germany) following DNAseI treatment and column purification according to the Direct-zol RNA Micro prep kit from ZymoResearch (R2062). The RNA integrity and quantity were measured with the Qubit fluorometer (Invitrogen, Q33216) and the 2100 Bioanalyzer (Agilent Technologies, Germany).

RNA-seq libraries were prepared from 300 ng of total RNA, using the Illumina TruSeq stranded mRNA library prep kit (Cat:20020594) and IDT for Illumina xGen UDI-UMI Adapters (Cat:10005903). Libraries were sequenced on an Illumina NextSeq500 (Mid output, 300 cyc) and NextSeq2000 (P3, 300 cyc) with 145 bp paired-end reads.

Nuclear genome assembly

Before assembly, PacBio HiFi reads were subjected to additional trimming to remove residual adapter sequences using NCBI’s VecScreen tool. Genome size was estimated using GenomeScope72 from Illumina short reads and PacBio HiFi data. The primary assembly was generated with hifiasm124 using a combination of HiFi and Hi-C reads.

Scaffolding was performed iteratively with YAHS77 on the phased haplotype 1. We adjusted the run parameters for scaffolding resolutions (-r 1000,2000,5000,10000,20000,30000,50000,70000,100000,150000,200000,250000,300000, 350000,400000,500000,700000,1000000,2000000,5000000,10000000,15000000,20000000,30000000,40000000,50000000,60000000,70000000,80000000,100000000,110000000,120 000000,150000000,170000000,200000000,500000000), repetitions per resolution (-R3), minimum read mapping quality (-q1) and telomeric sequence (--telo-motif “TTAGGG”) from the only experimentially determined telomere motif for cephalopods (Amphioctopus areolatus125, accessed via TeloBase126) to optimize the tool’s performance for a large, fragmented assembly. Finally, the assembly was manually curated using JBAT78 to place residual scaffolds into chromosome scaffolds. Different versions of the assembly were evaluated based on HiC coverage, mRNA alignment and completeness based on BUSCO75 v5.5.0 (metazoa_odb10 and mollusca_odb10).

Mitochondrial genome assembly

To assemble the mitochondrial genome, we aligned a published S. officinalis mitochondrial genome127 (NC_007895.1) to the PacBio Hi-Fi and Omni-C reads using minimap284. The hits were extracted using seqtk128 subseq and assembled using hifiasm124. The resulting contigs contained multiple repeats of the circular mitochondrial genome, and were aligned back to NC_007895.1 to extract the final sequence.

Nuclear genome annotation

Repetitive elements in the genome were softmasked using RepeatMasker v4.1.7-p189 (with rmblast v2.14.1+) after creating a custom repeat library with RepeatModeler v.2.0.688.

Gene models were created using BRAKER375,91–106, run via the Docker container, on the softmasked genome using both RNA-seq and protein data. We used long-read Iso-Seq data and short-read Illumina RNA-seq data generated for this study (see above). For protein data, publicly available peptide annotations for Doryteuthis pealeii46, Euprymna scolopes107, Octopus bimaculoides46, Octopus vulgaris43, Nautilus pompilius108, Pecten maximus51 were used for training, as well as a previously generated proteome for Sepia officinalis from StringTie gene models (this study). RNA-seq data was input into BRAKER3 using the —bam option, protein files were specified with the–prot_seq option. The BUSCO completeness of the resulting gene set was maximized using TSEBRA within BRAKER on the BUSCO lineage metazoa_odb10.

For the initial protein set used as input for BRAKER, the short and long RNA reads were aligned to the genome using minimap284. Transcript models were predicted with StringTie v3.0.0129 using the --conservative and --mix options. The resulting GTF files were combined using the transcript merge mode, resulting in a set of non-redundant transcripts. Finally, TransDecoder130 v5.7.0 was run with default parameters to translate coding regions in the transcripts.

The final gene annotations for S. officinalis were assessed for completeness using OMArk v.0.3.0109 with the ancestral clade Lophotrochozoa and without including splice information, accessed via their webserver. Proteins were annotated with orthology information using InterProScan110 v5.73-104 including lookup of annotations and GO terms (options -iprlookup -goterms). Further orthology information was added using the eggNOG-mapper v2.1.12111 webserver with the eggNOG v5.0 database and default parameters.

Whole genome alignment and synteny analysis

For whole genome alignments, the assembly produced for Sepia officinalis in this study and the published assembly from the Darwin Tree of Life project (GCA_964300435.1) were aligned using minimap284 and visualized with a custom script in R v4.4.2131.

For synteny analyses, peptide and gtf files were downloaded for Doryteuthis pealeii and Euprymna scolopes. Annotation files for Sepia esculenta were recently generated by liftover annotation from Sepia pharaonis87 and downloaded from Zenodo86. Synteny analyses between all chromosomes of the compared species were performed using the R package GENESPACE v.1.2.3132. S. officinalis was used as the reference species for all synteny maps.

Coverage analysis

Read coverage across chromosomes was analyzed as described recently87. Short reads were aligned using STAR v2.7.11b133 to our chromosome-scale assembly. For Sepia esculenta, short reads (ERR12945500) and assembly (GCA_964036315.1) were downloaded from NCBI. The sequencing depth was calculated using mosdepth134 by a window size of 500,000 bp and normalized to the median coverage of the first chromosome.

eLife Assessment

This manuscript reports a high-quality genome assembly of the European cuttlefish, Sepia officinalis, a representative species of the Cephalopod lineage. The data are based on current best practices for sequencing and genome assembly, including PacBio HiFi long reads and Hi-C chromatin conformation capture; the analysis is currently in parts incomplete, as further analyses are required to confirm the correct chromosome number. This genome will be a useful resource for the community of researchers interested in cuttlefish biology and comparative genomics in general.

Reviewer #1 (Public review):

Summary:

This manuscript presents a high-quality, chromosome-level genome assembly of the European cuttlefish (Sepia officinalis), a representative species of the cephalopod lineage. Using state-of-the-art sequencing and scaffolding technologies -including PacBio HiFi long reads and Hi-C chromatin conformation capture - the authors deliver a genome assembly with exceptional contiguity and completeness, as evidenced by high BUSCO scores. This genome resource fills a significant gap in cephalopod genomics and offers a valuable foundation for studies in neurobiology, behavior, and evolutionary biology. However, there are several major aspects that need to be strengthened.

Major Revisions Recommended:

(1) Single-individual genome limitation

The genome assembly is based on a single individual, which appears to be male. While this approach is common in genome projects, it does not capture the full genetic diversity of the species. As S. officinalis exhibits a wide geographical range and possible population structure, future efforts (or discussion in this manuscript) should consider re-sequencing multiple individuals - of both sexes and from diverse geographic origins - to characterize population-level variation, sex-linked features, and structural polymorphisms.

(2) Limited experimental validation of chromosomal inferences

The study reports chromosome-scale scaffolding using Hi-C data and proposes a revised karyotype for S. officinalis. However, these inferences would be significantly strengthened by orthogonal validation methods. In particular, fluorescence in situ hybridization (FISH) or karyotyping from cytogenetic preparations would provide direct confirmation of chromosome number and structural arrangements. The reliance solely on Hi-C contact maps for inferring chromosomal organization should be acknowledged as a limitation or supplemented with such validations.

(3) Shallow discussion of chromosomal evolution

The manuscript briefly mentions chromosomal number differences among cephalopods but does not explore their evolutionary or functional implications. A more thorough comparative analysis - linking chromosomal rearrangements (e.g., fusions, fissions) with ecological adaptation, life history, or neural complexity - would greatly enhance the impact of the findings. Referencing chromosomal dynamics in related taxa and possible links to behavioral innovations would contextualize these results more effectively.

(4) Underdeveloped gene family and pathway analysis

While the authors identify expansions in gene families such as protocadherins and C2H2 zinc finger transcription factors, the functional significance of these expansions remains speculative. The manuscript would benefit from:

a) Functional enrichment analyses (e.g., GO, KEGG) targeting these gene families.

b) Expression profiling across tissues or developmental stages to infer regulatory roles.

c) Comparison with expression or expansion patterns in other cephalopods with known behavioral complexity (e.g., Octopus bimaculoides, Euprymna scolopes).

d) Potential integration of transcriptomic or epigenomic data to support regulatory hypotheses.

Reviewer #2 (Public review):

Summary:

This paper concerns an interesting organism, Sepia officinalis. However, in the opinion of this reviewer, the paper reads somewhat like a genome report. The authors have used 23x PacBio HiFi in conjunction with relatively low coverage (11x) Hi-C to scaffold the genome into a karyotype of 47 chromosomes. They have used a combination of short and long read RNA seq to annotate the genome in what looks like a very good annotation. The paper offers basic analyses of the Busco evaluation, some descriptive analyses of gene family and repeat content, and a bit more focused analysis on synteny among sequenced squids. Generally, the data will be useful.

Strengths:

This is a high-quality annotation, and the data ultimately will be useful to other researchers. I appreciate trying to understand what's happening between assemblies of S. officinalis.

Weaknesses:

I don't believe the data at hand makes a strong case for the argument of 47 chromosomes. This is my biggest sticking point with the paper, and it is for a few reasons:

(1) The authors point to assembly differences between the DToL assembly and the one presented in the manuscript and seem to claim that DToL is incorrect. However, the DToL assembly (xcSepOffi3.1) is based on much deeper HiFi and HiC coverage than the one at hand (51x and 80+x respectively). There are many things to try here, including:

a) Downloading the DToL data and reassembling using a common pipeline.

b) Downsampling the DToL data to similar coverage as what the authors have achieved.

c) Combining your data and that of DToL for even deeper coverage (heterozygosity is low enough that I don't imagine this impeding things too badly).

(2) Looking at Figure 1, there appears to be a misjoin at chromosome 42. Looking carefully at Figure S1, that misjoin does not appear on any of the panels - this is confusing. Given the size of that chromosome and the authors' chromosome numbering, I'm guessing this is a manual merge (as it's larger than most of the chromosomes numerically close (40, 41, 43, etc). Further, staring closely at Figure 1, there appear to be cross-scaffold contacts between 42 and 43 and 42 and 44. Secondarily there are contacts between 43 and 44. This bit of the assembly seems potentially problematic.

Reviewer #3 (Public review):

Summary:

In this study, authors Simone Rencken and co-authors present and investigate the genome of the common cuttlefish Sepia officinalis.

Strengths:

The authors explain in a detailed yet concise manner the main steps for a genome assembly, with very robust methods for validation, and according to current best practices. In addition to the chromosomal assembly, the authors confirmed the presence of 47 chromosomes using Hi-C data and multiple species synteny. They also generated a comprehensive gene annotation, with assessments of gene completeness, providing a useful resource for the community of researchers interested in cuttlefish biology and comparative genomics.

Weaknesses:

While the study touches upon the subjects of gene content, TE activity, or species-level comparisons, the study does not provide in-depth investigations of these.

Author response:

The following is the authors’ response to the original reviews.

Public Reviews:

Reviewer 1 (Public review):

Summary:

This manuscript presents a high-quality, chromosome-level genome assembly of the European cuttlefish (Sepia officinalis), a representative species of the cephalopod lineage. Using state-of-the-art sequencing and scaffolding technologies -including PacBio HiFi long reads and Hi-C chromatin conformation capture - the authors deliver a genome assembly with exceptional contiguity and completeness, as evidenced by high BUSCO scores. This genome resource fills a significant gap in cephalopod genomics and offers a valuable foundation for studies in neurobiology, behavior, and evolutionary biology. However, there are several major aspects that need to be strengthened.

Major Revisions Recommended:

(1) Single-individual genome limitation

The genome assembly is based on a single individual, which appears to be male. While this approach is common in genome projects, it does not capture the full genetic diversity of the species. As S. officinalis exhibits a wide geographical range and possible population structure, future efforts (or discussion in this manuscript) should consider re-sequencing multiple individuals - of both sexes and from diverse geographic origins - to characterize population-level variation, sex-linked features, and structural polymorphisms.

We thank the reviewer for this summary and the important point raised. While sequencing additional individuals, unfortunately, lies outside the scope of our study, we used the published data from the DToL assembly (from a male individual from a different geographical origin) to begin to investigate their differences.

First, we attempted to create a mixed assembly from both datasets, as also suggested by Reviewer 2, to increase data coverage and genetic information. Even though the heterozygosity estimate is quite low (ca. 1%), the mixed assembly produced severely inflated and fragmented results, yielding an assembly ca. 3× larger than expected, with the top 46 contigs covering only ~5% of the total length - a sign of over duplication and failed haplotype collapse.

This result is not surprising when considering the assembly algorithms: most programs, including hifiasm used in this study, assume a single diploid individual (or a trio assembly including data from both parents), so using multiple individuals breaks this assumption. Assembly pipelines infer homozygous/heterozygous coverage cutoffs from the k-mer histogram. Mixing individuals raises apparent heterozygosity far above true diploid levels, turning the expected bimodal k-mer profile into a complex multimodal distribution. This misleads the phasing and purging steps in the assembly pipeline, causing over-expansion and fragmentation of the assembly.

Second, we created separate assemblies from the raw data sets of MPIBR and DToL using the exact same pipeline and parameters to avoid the technical problem described above. These assemblies are directly comparable, and after aligning them, it is possible to build a pangenome graph that we believe would help to address the points raised by the reviewer. Pangenome graphs can represent cross-individual variation more accurately and improve read alignment in regions of high genomic variation, which can aid population-level analyses [1]. We agree on the importance of this work, yet collecting data from more individuals and the construction and analysis of a pangenome graph lies beyond the scope of this manuscript and should be part of future efforts by the cephalopod genomics field.

(2) Limited experimental validation of chromosomal inferences

The study reports chromosome-scale scaffolding using Hi-C data and proposes a revised karyotype for S. officinalis. However, these inferences would be significantly strengthened by orthogonal validation methods. In particular, fluorescence in situ hybridization (FISH) or karyotyping from cytogenetic preparations would provide direct confirmation of chromosome number and structural arrangements. The reliance solely on Hi-C contact maps for inferring chromosomal organization should be acknowledged as a limitation or supplemented with such validations.

We appreciate the reviewer’s point regarding the value of orthogonal validation methods to support the chromosome-scale scaffolding and proposed karyotype. We acknowledge that relying solely on Hi-C contact maps to infer chromosome number and structure presents limitations, as also becomes apparent in our detailed analysis of both S. officinalis genome assemblies (in Figure 2 and Supplementary Figure 3 of the revised manuscript). We attempted to complement these analyses with cytogenetic approaches. Unfortunately, the availability of suitable mitotic tissue was limited. Moreover, our karyotyping trials proved challenging: resolving the ≥92 (2n) chromosomes in situ was not feasible due to their high number and the small size of the nuclei (approximately 5 µm in diameter on average).

We now highlight this point as an important direction for future work in our discussion (line 456-466):

“Additional methods such as cytogenetic karyotyping or optical mapping such as BioNano [141] (imaging of fluorescently tagged, linearized DNA) could be used to validate chromosome numbers. However, whereas karyotypes of octopuses have been consistent throughout the literature (1n=30) [142,143], those measured in decapods vary greatly. For example, 1n=46 chromosomes have been reported for two species of cuttlefish (A. esculentum and A. lycidas) and three loliginid squids [85]; 1n=36 has been reported for A. Arabica [86] and 1n=24 in A. pharaonis [87]. In S. officinalis, a karyotype of 1n=52 is reported for testis samples [88]. Combining cytogenetic preparations with fluorescent labeling of centromeric or telomeric sequences, as demonstrated in the octopus A. aerolatus [143] could help resolve these issues. Establishing a routine staining protocol would enable comprehensive tests at the species- and population-level.”

(3) Shallow discussion of chromosomal evolution

The manuscript briefly mentions chromosomal number differences among cephalopods but does not explore their evolutionary or functional implications. A more thorough comparative analysis - linking chromosomal rearrangements (e.g., fusions, fissions) with ecological adaptation, life history, or neural complexity - would greatly enhance the impact of the findings. Referencing chromosomal dynamics in related taxa and possible links to behavioral innovations would contextualize these results more effectively.

We agree with the reviewer that this is a fascinating topic of research that demands further attention and have extended our discussion, which now reads (line 476-501):

“In addition to studying chromosomal topology in phylogenetic reconstructions, some of the most interesting aspects of these rearrangements relate to changes of and innovation in regulatory elements that underlie phenotypic diversity. In coleoid cephalopods, it is thought that an ancient large-scale genome rearrangement was combined with lineage-specific changes and repeat expansions [48–50]. This restructuring gave rise to hundreds of tightly linked, evolutionarily unique microsyntenies, corresponding to distinct topological compartments with specialized regulatory architectures that contribute to complex, tissue-specific expression patterns in the nervous system and elsewhere [43]. Extending this, chromosomal conformation analyses in E. scolopes revealed that co-regulated eye and light-organ genes cluster at topologically associating domain (TAD) boundaries, and that an evolutionarily recent rearrangement at the dachshund (DAC) locus may have been instrumental in the emergence of the symbiotic light organ in Euprymna - directly linking specific chromosomal topology to morphological innovation [44].

To understand the broader functional impact of these changes across coleoids, a recent study investigating Micro-C, RNA-seq, and ATAC-seq data from multiple species revealed broadly conserved chromatin domains, but also many lineage-specific chromatin loops that form novel regulatory signatures and impact expression profiles across species and tissues [149].

Despite the observed small-scale regulatory changes, the chromosomes of decapods are considered to be more closely related to the ancestral coleoid karyotype than those of octopods. The derived octopod karyotype becomes apparent when comparing it to the genome of the vampire squid, an early-branching octopodiform (sister to all octopods) which retained features of the decapod, ancestral karyotype [150]. Taken together, the conserved karyotype of decapods accommodates fine-scale regulatory diversity that might underlie morphological diversity among species, which suggests that many regulatory innovations are still being evolutionarily explored through rearrangements within the existing chromosomes.”

(4) Underdeveloped gene family and pathway analysis

While the authors identify expansions in gene families such as protocadherins and C2H2 zinc finger transcription factors, the functional significance of these expansions remains speculative. The manuscript would benefit from:

(a) Functional enrichment analyses (e.g., GO, KEGG) targeting these gene families.

(b) Expression profiling across tissues or developmental stages to infer regulatory roles.

(c) Comparison with expression or expansion patterns in other cephalopods with known behavioral complexity (e.g., Octopus bimaculoides, Euprymna scolopes).

(d) Potential integration of transcriptomic or epigenomic data to support regulatory hypotheses.

We thank the reviewer for these constructive suggestions and have substantially expanded the functional characterization of expanded gene families in the revised manuscript.

To address points a) + b), we performed GO enrichment analyses for all expanded gene families (orthogroups), both for the largest gene families and the most significantly expanded families identified from our CAFE5 analysis. Further, we cross-referenced all S. officinalis members of each expanded orthogroup against differentially expressed genes in our bulk RNA-seq data from multiple tissues (initially collected to improve the gene modeling), allowing us to infer tissue-specific expression patterns for the expanded families.

To address point (c), the species-resolved copy-number profiles from our orthogroup analysis directly situate the S. officinalis expansions within the broader coleoid context, including O. bimaculoides, O. vulgaris, E. scolopes, and D. pealeii, enabling direct comparison of expansion scale and lineage specificity across species with varying degrees of behavioural complexity. We note that the C2H2 zinc finger and protocadherin expansions show distinct phylogenetic profiles consistent with independent radiations in octopods and decapodiforms, in agreement with recent studies.

Regarding point (d), no epigenomic data for S. officinalis was publicly available at the time of writing, thus we focused on the transcriptomic data from this study, as described above.

We describe this analysis in two additional results paragraphs to the manuscript, one modified (Figure 4) and two new figures (Figure 5 and Supplementary Figure 7), which are reproduced (lines 294-400):

“Analysis of expanded gene families

We sought to investigate the S. officinalis gene annotation and place it in the context of gene repertoires from other cephalopod or molluscan species. First, we collected available genome annotations from 12 other molluscan species (Table 2) and clustered them using OrthoFinder v.3.1.0 [122], resulting in 23,658 orthogroups, hereafter named gene families.

First, we investigated 36 of the gene families that contain more than 100 genes in any of the species, with 17 of these families containing at least one gene of S. officinalis, that reflect large-scale gene family expansions (Figure 4E). We used the InterProScan and eggNOG-mapper annotations to infer functional roles of these genes, selecting the most common gene annotation as the name of the gene family.

The zinc finger C2H2-type transcription factors (TFs) were grouped into three of the large gene families, with the largest family (OG0000000) only present in decapod cephalopods. This likely reflects the largely independent expansions in the octopod and decapod lineages that date back to a burst of transposon activity ca. 25 million years ago [46,48,49]. The largest expansion across mollusks occurs in the cadherin-like family (OG0000001): 310 in S. officinalis, 283 in D. pealeii, 209 in A. lycidas, 102 in O. vulgaris, 55 in O. bimaculoides, with low but non-zero counts in bivalves (C. virginica, M. gigas). This profile is consistent with the protocadherin expansion first described in O. bimaculoides [46] and subsequently shown to be present across cephalopods [48,49,123].

HPGDS (OG0000005, hematopoietic prostaglandin D synthase) is a glutathione-S-transferase family member that catalyzes the conversion of prostaglandins, which have well-described roles in immune responses in vertebrates and insects [124,125]. This family shows a broad expansion in decapods, with a lesser expansion in octopods. Additionally, members of the glutathione-S-transferase families have been co-opted as S-crystallins, structural proteins found in the lens of cephalopods that may, or may not, retain enzymatic functions [126,127].

Two large families are mostly lineage-restricted. The RING-type zinc finger family (OG0000058) has 103 copies in S. officinalis and 26 in A. lycidas but is absent in all other species except for E. scolopes. Conversely, OG0000002 (unknown function) has 479 copies in E. scolopes and only a few copies in the other species. This interesting Sepiolid-specific expansion warrants further characterization.

We estimated gene family evolution rates using CAFE5 [128] for all families with less than 100 copies in any species (this excludes the families described above, as very large copy-number differences between species preclude likelihood calculations under the applied birth-death model). After comparing different model parameters, we chose a gamma model with three rate categories, allowing for evolutionary rate variation among gene families. Out of the 12,895 gene families analyzed, 1,813 showed a significant (p < 0.05) expansion or contraction in at least one of the species. We focused our analysis on the 30 most significantly expanded families; among them were several retrotransposon-associated domains that have expanded specifically in S. officinalis five families carrying Retrovirus-related Pol polyprotein domains, two Reverse transcriptase domain families, and four Ribonuclease H-like families (Supplementary Figure 7A). There was no coordinate-based overlap of the coding sequences with annotated TEs from the RepeatMasker output (Methods).

In addition to the three large gene families of C2H2 zinc finger expansions, 45 gene families containing this TF type showed a significant change in the CAFE5 analysis. Notably, eight of the significant gene families, as well as four of the largest gene families, were annotated as CCHC-type zinc fingers, which contain a “zinc knuckle” motif that is characteristic of retroviral nucleocapsid proteins [129] and is functionally integrated in the genomes of several species, including humans [130].

Some gene families without any relationship to retrotransposons were also expanded. For example, the UGT2A1-related family is a UDP-glucuronosyltransferase, a class of enzymes central to phase II detoxification and conjugation of metabolites, reported in other mollusks in the context of environmental chemical tolerance [131], and in insects in the context of pigmentation [132]. We also detected a family of homeodomain-like proteins, representing an expansion of this important TF family.

Tissue-specific expression of expanded gene families

To place the identified gene families in a functional context, we profiled their expression in the bulk RNA-seq data (taken from multiple tissues of S. officinalis) used originally for gene modeling (Figure 5A). Principal component analysis (PCA) revealed the largest axis of variation in gene expression to separate brain tissues from peripheral tissues, with skin being the most transcriptomically distinct (Figure 5A), consistent with the high number of tissue-specific differentially expressed (DE) genes identified in non-neural tissues (Figure 5B). We identified the genes belonging to expanded families that were differentially expressed across tissues and enriched gene ontology [133,134] (GO) terms for them to gain additional insight. The large families excluded from CAFE5 modelling and the significantly expanded families identified by CAFE5 were analyzed separately.

Eleven of the largest gene families were expressed in our data (Figure 5C) and five had enriched GO terms (Figure 5D,E). Among them, the cadherin family showed brain-restricted expression and GO terms related to cell–cell adhesion and calcium binding, consistent with their role in neuronal connectivity and circuit formation [46,135]. Two C2H2 zinc finger gene families were expressed in the optic and vertical/subvertical lobes of the brain and in the skin, with GO terms related to DNA-binding, transcriptional regulation or development. The RING-type zinc finger family was expressed specifically in the skin, with GO terms including zinc binding and ubiquitin protein ligase activity, the canonical function of RING-domain E3 ligases [136]. Genes of the HPGDS/S-crystallin family were expressed in the brain (basal and optic lobes and posterior subesophageal mass) and skin, with GO terms related to glutathione metabolism, matching their described enzymatic function. We did not find expression in the retina, which is expected given that S-crystallins are expressed in lentigenic cells of the eye [42,137] and these cells were not included during sampling.

Among the 30 most significantly expanded families examined (out of 1,813 total), expression was widespread (20/30) and tissue-specific differential expression was common (17/30), suggesting that a substantial proportion of expanded paralogs represent functional coding sequences with specialized spatial deployment (Supplementary Figure 7B). Ten of the retrotransposon-associated families were differentially expressed in the brain (optic and vertical/subvertical lobes) and skin, arguing against these loci being inactive repeat fragments and supporting their inclusion as transcribed gene models. Two significantly expanded families showed both differential expression and enriched GO terms (Supplementary Figure 7C). The first was the UGT2A1-related family, which had the largest number of differentially expressed genes overall, with expression concentrated in the skin, retina and posterior subesophageal mass of the brain. Enriched GO terms matched the described enzymatic function for this family, namely UDP-glycosyltransferase activity. The second gene family was the homeodomain-like family with enrichment for DNA binding terms consistent with their role as transcription factors, and was preferentially expressed in the vertical and subvertical brain lobes with weaker expression in other areas.

Collectively, many differentially expressed genes from expanded families were restricted to specific tissues or brain subregions (Figure 5F and Supplementary Figure 7D), indicating that paralogs within an expanded family have adopted distinct spatial expression domains and possibly, specialized functions.”

Reviewer 2 (Public review):

Summary:

This paper concerns an interesting organism, Sepia officinalis. However, in the opinion of this reviewer, the paper reads somewhat like a genome report. The authors have used 23x PacBio HiFi in conjunction with relatively low coverage (11x) Hi-C to scaffold the genome into a karyotype of 47 chromosomes. They have used a combination of short and long read RNA seq to annotate the genome in what looks like a very good annotation. The paper offers basic analyses of the Busco evaluation, some descriptive analyses of gene family and repeat content, and a bit more focused analysis on synteny among sequenced squids. Generally, the data will be useful.

Strengths:

This is a high-quality annotation, and the data ultimately will be useful to other researchers. I appreciate trying to understand what's happening between assemblies of S. officinalis.

Weaknesses:

I don't believe the data at hand makes a strong case for the argument of 47 chromosomes. This is my biggest sticking point with the paper, and it is for a few reasons:

(1) The authors point to assembly differences between the DToL assembly and the one presented in the manuscript and seem to claim that DToL is incorrect. However, the DToL assembly (xcSepOffi3.1) is based on much deeper HiFi and HiC coverage than the one at hand (51x and 80+x respectively). There are many things to try here, including:

(a) Downloading the DToL data and reassembling using a common pipeline.

(b) Downsampling the DToL data to similar coverage as what the authors have achieved.

(c) Combining your data and that of DToL for even deeper coverage (heterozygosity is low enough that I don't imagine this impeding things too badly).

We thank the reviewer for these helpful suggestions and want to clarify that we did not seek to point out errors in the DToL assembly, but rather to investigate the unexpected discrepancies between the two assemblies. It is correct that the DToL data has a much higher coverage than our data. We followed the individual suggestions and incorporated them into the revised manuscript. We reproduce the relevant sections below, and provide additional information:

(a) Downloading the DToL data and reassembling using a common pipeline.

We downloaded the DToL data and reassembled it using a common pipeline, yielding the results listed in Author response table 1. The DToL assembly is more contiguous, which is mainly due to its higher HiFi coverage. It also receives slightly better BUSCO scores (computed using odb12 as recommended by Reviewer 3).

Full statistics of S. officinalis assemblies from two independent datasets, assembled using a common pipeline.

The updated manuscript now reads (lines 146-159):

“A chromosome-scale assembly for Sepia officinalis was released recently by the Wellcome Sanger Institute’s Darwin Tree of Life project [75] (DToL, GCA_964300435.1). That genome was assembled from a male individual using high coverage PacBio Sequel II (~51x) and Arima2 Hi-C (~80x) data, with a final assembly size of 5.8 Gb. The the haploid chromosome number was estimated to be 49. To compare both S. officinalis datasets directly, we downloaded the DToL data and created two new assemblies using the pipeline described above (hifiasm using PacBio HiFi and Hi-C data). The resulting assemblies were overall very similar, with the DToL assembly having a slightly higher contiguity (N50 length, see Table 1) and BUSCO completeness (Supplementary Figure 2A,B) due to their higher sequencing coverage.”

To further compare the two datasets, we added a new Figure 2 to the revised manuscript and the following paragraph to the results (lines 160-169):

“After scaffolding with YAHS, both datasets reached the previously identified chromosome numbers (1n=47 for MPIBR and 1n=49 for DToL, Figure 2A,B). To further investigate this surprising discrepancy, we aligned both assemblies using Winnowmap [89] to locate the differences between them (Figure 2C). We observed four “breakpoints” (BP) of chromosome scaffolds: one in the MPIBR assembly compared to DToL (BP1: DToL_5 = MPIBR_40+44) and three in the DToL assembly compared to MPIBR (BP2: DToL_31+40 = MPIBR_2, BP3: DToL_41+46 = MPIBR_6, BP4: DToL_44+45 = MPIBR_7). We also aligned the assemblies to the chromosome-scale genome of another cuttlefish Acanthosepion esculentum (1n=46, GCA_964036315.1). In this alignment, all four breakpoints were collinear with single A. esculentum chromosomes (Figure 2D).”

(b) Downsampling the DToL data to similar coverage as what the authors have achieved.

Instead of downsampling the DToL data, we decided to analyze the Hi-C and HiFi data for both assemblies, focusing on the four “breakpoints” between the assemblies and the A. esculentum genome that we described above. First, we performed a QC analysis of the Hi-C reads using pairtools [2], the result is visualized in Author response image 1. The percentage of valid Hi-C read pairs, i.e., cis pairs with insert distances of more than 1 kb and trans pairs, following the Dovetail genomics QC manual (https://dovetail-analysis.readthedocs.io/en/latest/whole_genome/qc.html). When Hi-C pairs were aligned to the primary contigs from hifiasm (as is used for scaffolding with YAHS), the DToL HiC data contains fewer valid read pairs (11.4%) than the MPIBR data (43.1%), possibly due to using a different tissue (eye vs. optic lobe) and HiC kit (Arima 2 vs. Dovetail OmniC) for the library preparation. Nonetheless, due to the much higher overall coverage, the amount of valid read pairs is still 2.35x higher for DToL (144,014,368 pairs) than for MPIBR (61,318,955 pairs). The higher trans fraction (i.e. HiC pairs across contigs) is dependent on the length of the primary contigs, so the higher trans fraction for the MPIBR data can be explained by the lower contiguity of its primary contigs. It is conceivable that for both assemblies, the low numbers of valid read pairs introduce a technical fragmentation of certain chromosomes, as indicated by the identified breakpoints (Figure 2).

Analysis of Hi-C read pairs from both S. officinalis assemblies.

Hi-C reads were aligned to the primary contigs from hifiasm (as is used for scaffolding with YAHS) and analyzed using pairtools. Note the higher fraction of long-range contacts (at least 1 kb cis pairs or trans pairs) in the MPIBR data (top) compared to DToL (bottom). Due to overall higher coverage, the absolute number of read pairs is higher for DToL than for MPIBR data.

Second, we performed a detailed analysis of read coverage along the breakpoint junctions of the discrepant chromosomes/scaffolds between both assemblies. We included a description of the results and a new Supplementary Figure 3 in the manuscript, (lines 171-207):

“To better understand the potential cause of these divergent chromosome numbers, we analyzed the Hi-C and HiFi coverage in the breakpoint regions (Supplementary Figure 3A). First, we aligned the Hi-Fi reads to the scaffolds and extracted all alignments along the 200 kb terminal scaffold windows to find any notable drops in coverage, or reads spanning any of the scaffold junctions. We detected no spanning reads. This is not surprising given that no contigs were assembled at these sites, resulting in the observed scaffold junctions. More interestingly, we noted a ~5-fold decrease in HiFi coverage along the DToL scaffold_40 (part of BP2) relative to its flanking regions, indicating a highly repetitive, low-mappability region at this boundary.

Next, we realigned the Hi-C data to the scaffolded assemblies using bwa-mem2 [91] and extracted all trans HiC pairs (between-scaffold contacts) using pairtools [92]. We normalized trans HiC contacts to the scaffold length and compared contact rates between breakpoint scaffolds to the baseline contact rate (computed from pairs of scaffolds with a clear 1-to-1 match between assemblies), and the contact rate within scaffolds (intra-scaffold pairs) (Supplementary Figure 3B,C). The contact rates within breakpoints were consistently lower than within scaffolds, likely falling below the threshold to be merged during assembly. However, the contact rates at three of four breakpoints (BP1, BP3, BP4) were significantly elevated above the genome-wide background distribution (empirical p = 0.010, 0.005, 0.005 respectively), suggesting that they may represent intra-chromosomal contacts disrupted by a misassembly. Notably, BP2 was not significant (empirical p = 0.170), likely due to the low coverage and mappability around the DToL scaffold_40 boundary. Considered jointly, the three DToL breakpoint scaffold pairs showed significantly higher trans contact rates than the background (Wilcoxon rank-sum, one-tailed, U = 1771, p = 0.004).

Lastly, we analyzed the repeat landscape around the 200 kb scaffold ends using RepeatMasker [93] and the custom repeat library that we had generated for Sepia officinalis (described further below). Compared to control scaffolds of the same assembly, we observed consistently elevated repeat content at the breakpoint junctions (mean 71.5% vs 67.6% masked bases), with an enrichment of unclassified repeats (32.1% vs 30.0%), which could explain a repeat-driven assembly fragmentation or scaffolding failure. The BP2 DToL scaffold_40 junction window was 99.99% masked (99.2% unclassified repeats), providing a likely mechanistic explanation for both the HiFi coverage drop and the absence of a significant trans Hi-C signal at this breakpoint. Taken together, these analyses suggest that the different chromosome numbers across the two S. officinalis assemblies are due to technical reasons, caused by repeat-rich scaffold boundaries that impair HiFi and Hi-C read alignment and in turn, correct assembly in these regions.”

(c) Combining your data and that of DToL for even deeper coverage (heterozygosity is low enough that I don't imagine this impeding things too badly).

When combining the data to achieve a higher coverage, we ran into the assembly fragmentation issues detailed above in response 1) to Reviewer 1.

(2) Looking at Figure 1, there appears to be a misjoin at chromosome 42. Looking carefully at Figure S1, that misjoin does not appear on any of the panels - this is confusing. Given the size of that chromosome and the authors' chromosome numbering, I'm guessing this is a manual merge (as it's larger than most of the chromosomes numerically close (40, 41, 43, etc). Further, staring closely at Figure 1, there appear to be cross-scaffold contacts between 42 and 43 and 42 and 44. Secondarily there are contacts between 43 and 44. This bit of the assembly seems potentially problematic.

This is a great observation, indeed the HiC maps differ between Figure 1 and Figure S1. Figure 1 is the result of scaffolding with YAHS and manual curation, whereas Figure S1 was scaffolded using HapHiC. We updated the figure legend to clarify this important difference. HapHiC produces very clean contact maps without the need for manual curation, but when analyzed at a higher resolution, the tool broke many contigs and ultimately compromised the assembly quality, possibly due to our comparatively low HiC coverage. Thus, we preferred to use YAHS and manual curation, which is perhaps inherently error-prone, as becomes apparent in the regions of the assembly that are pointed out by the reviewer.

Reviewer 3 (Public review):

Summary:

In this study, authors Simone Rencken and co-authors present and investigate the genome of the common cuttlefish Sepia officinalis.

Strengths:

The authors explain in a detailed yet concise manner the main steps for a genome assembly, with very robust methods for validation, and according to current best practices. In addition to the chromosomal assembly, the authors confirmed the presence of 47 chromosomes using Hi-C data and multiple species synteny. They also generated a comprehensive gene annotation, with assessments of gene completeness, providing a useful resource for the community of researchers interested in cuttlefish biology and comparative genomics.

Weaknesses:

While the study touches upon the subjects of gene content, TE activity, or species-level comparisons, the study does not provide in-depth investigations of these.

We thank the reviewer for their positive assessment of our manuscript. We acknowledge the descriptive nature and limitations of our previous analyses of gene content, TE distribution, and species comparisons. Our focus for the initial submission was to provide a high-quality assembly that could serve as a resource for anyone interested in Sepia officinalis or related species. However, we agree that greater insight into genome content is valuable as well. In the revised manuscript, we included a more detailed analysis of expanded gene families and GO enrichment analysis of our bulkRNAseq data, which we summarized in response 4) to reviewer 1.

Recommendations for the authors:

Reviewer #1 (Recommendations for the authors):

Minor Revisions Recommended:

(1) Figure and legend clarity

Several figures lack sufficient annotation. All figures, including supplementary ones, should include:

(a) Clear axis labels.

(b) Descriptions of statistical measures (n values, error bars, statistical tests).

(c) Legends that allow the figure to be understood independently of the main text.

We updated the figures accordingly.

(2) Terminology and formatting

(a) Consistency in gene and species nomenclature should be maintained throughout (e.g., italicizing gene names and Latin binomials).

(b) Ensure that abbreviations (e.g., Hi-C, BUSCO, FISH) are defined upon first use.

We updated the nomenclature throughout the text and checked the definition of abbreviations used in the text. Further, we updated the names of several cuttlefish species according to the recent revision of genera, e.g. Sepia esculenta was changed to Acanthosepion esculentum [3].

(3) Literature coverage

The references primarily focus on earlier studies from 2010-2020. It would strengthen the context to include recent high-impact studies on cephalopod genomics and chromosomal biology published in the last 3 years (e.g., 2022-2024).

We apologize for this oversight and have extended the manuscript to discuss more of these recent studies.

(4) Clarify methods

While the methods section is generally detailed, some critical aspects are underspecified:

(a) Parameters used in genome annotation tools (e.g., BRAKER, RepeatMasker).

We thank the reviewer for bringing our attention to this shortcoming, and have added the missing parameters to the methods section. Additionally, the full code is available at https://gitlab.mpcdf.mpg.de/mpibr/laur/cuttlefishomics/soffgenome

(b) Criteria for ortholog clustering and gene family expansion analysis.

The details have been added to the methods section, which now reads (lines 828-853):

“Orthogroups were inferred across 13 molluscan species (Table 2), including S. officinalis, using OrthoFinder v3.1.0 [122] with default parameters. The input proteomes included the longest protein isoform per gene for each species. The rooted species tree from OrthoFinder [182,184] was converted to an ultrametric tree using the R package ape [183] v5.8.1.

Gene families were filtered by removing orthogroups present in only a single species, and by separating orthogroups containing 100 or more gene copies in any species, as extreme copy-number differences in gene families prevent likelihood calculation under the applied birth-death model.

Gene family evolution rates were estimated using CAFE5 [128] v5.1.1 on the filtered orthogroups, using the ultrametric species tree as input. Four models were evaluated: the base model (single global lambda), and Gamma models with k = 2, 3, and 4 rate categories, which allow evolutionary rate variation among gene families. The Gamma k = 3 model was selected based on the best (lowest) final log-likelihood score. All subsequent statistical inferences were performed under this model.

For families showing statistically significant expansion or contraction (p < 0.05 after Bonferroni correction), branch-specific copy-number changes were extracted from the CAFE5 output. Families were categorized as S. officinalis-specific, coleoid-specific, or broad expansions based on the distribution of significant changes across the phylogeny.

To assess whether expanded gene families in S. officinalis contained genes derived from or embedded within repetitive elements, a coordinate-based overlap analysis was performed. For each gene in an expanded orthogroup, the overlap between its coding sequence (CDS) coordinates and RepeatMasker annotations was computed using bedtools intersect v2.30 [185]. To avoid double-counting when multiple repeat annotations overlapped the same coding bases, overlapping repeat intervals were merged per gene prior to summing covered bases, and the overlap fraction was computed as merged covered bases divided by total CDS length.”

(c) Thresholds or cutoffs for synteny or duplication detection.

We included the details in the updated methods (lines 755-781):

“Synteny analyses between all chromosomes of the compared species were performed using the R package GENESPACE v.1.2.3 [175] with default parameters, described briefly below. Protein sequence similarity was first estimated using DIAMOND2 [109] in fast mode, and orthogroups and pairwise orthologues were inferred using OrthoFinder v2.5 [176] with hierarchical orthogroups (HOGs) enabled. Prior to synteny inference, tandem arrays were condensed to their most central representative gene, and gene rank order was recalculated on these array-representative genes to reduce confounding effects of tandem duplication on collinearity detection.

Syntenic blocks were identified pairwise between all genome combinations using MCScanX [177], constrained to DIAMOND hits where both query and target genes belonged to the same orthogroup (onlyOgAnchors = TRUE). Initial anchor hits were clustered into large syntenic regions using a density-based spatial clustering approach (dbscan [178]), with a minimum block size of five anchor genes (blkSize = 5) and a maximum of five intervening non-anchor genes permitted within a block (nGaps = 5). Anchor clustering used a search radius of 25 gene-rank positions (blkRadius = 25). All hits falling within a syntenic buffer of 100 gene-rank positions around confirmed block anchors (synBuff = 100) were retained as syntenic. No secondary syntenic hits were included (nSecondaryHits = 0). Syntenic orthogroups were integrated across all pairwise comparisons and collapsed into a pan-genome annotation anchored to. S. officinalis was used as the reference genome.

Syntenic relationships were visualized as riparian plots and pairwise dotplots using the built-in plotting functions of GENESPACE v1.2.3. Riparian plots were constructed using physical chromosomal coordinates (useOrder = FALSE) with S. officinalis as the reference, displaying all three genomes. A second riparian plot was generated highlighting a region of interest. Pairwise dotplots were produced species for the S. officinalis–D. pealeii and S. officinalis–E. scolopes genome comparisons, displaying only synteny-validated hits (type = "syntenic") with a minimum synteny score of 10 (minScore = 10) and a minimum of 10 genes per chromosome pair required for display (minGenes2plot = 10).”

Reviewer #2 (Recommendations for the authors):

Line 153 should be supplemental Figure 3B.

The text was referring to the correct Figure 2B (three species synteny comparison). It is now updated to Figure 3B in the revised manuscript.

Reviewer #3 (Recommendations for the authors):

(1) L37: Perhaps add a comparison with other species (mammals, Drosophila, etc.) to put this number in context.

We agree with this recommendation and added numbers for Drosophila and mouse to the text (lines 40-45):

“Coleoid cephalopods (octopus, squid, cuttlefish) are a highly derived group of mollusks, characterized by the largest nervous systems among all invertebrates (ca. 500 million neurons in an adult octopus of which 200 million are in the central brain [1,2], compared to ca. 140,000 in the fruit fly [3] or 70 million in the mouse [4]) and specializations with a great historical importance for neuroscience (e.g., “giant axons” [5] and “giant synapses” [6–8]).”

(2) L51, 279: "Octopodiformes" is a superorder, not a genus or a species name. It should not go in italics.

We updated this throughout the text.

(3) L53: "even smaller" seems odd here, because the argument of the sentence is to stress the large genome size of Octopodiformes. Perhaps start the sentence by stating that it is sometimes smaller, but often larger.

We rephrased the sentence for clarity, it now reads (lines 55-58):

“While the genomes of Octopodiformes (Octopus, Eledone, Argonauta) are either smaller than (1.1 Gigabases or Gb [45]) or comparable in size to that of humans (around 3 Gb [46,47]) the typical genomes of Decapodiformes (squids and cuttlefish) often reach 6 Gb [48,49].”

(4) L90: What tool was used to estimate the k-mer distribution of the long reads? Jellyfish? FastK? It's not mentioned anywhere in the text.

(5) L95: What k-mer size did the authors use to estimate k-mer distribution?

We thank the reviewer for pointing out this missing information, and have included the details in the methods (lines 692-694):

“The k-mer distribution was estimated using Meryl [165] within the Merfin [166] package with a k-mer size of 21, and genomeGenome size was estimated using GenomeScope [77] from Illumina short reads and PacBio HiFi data.”

(6) L99: What about using the most recent BUSCO databases? odb12?

We thank the reviewer for this question, which prompted us to compute BUSCO scores using the more recent odb12 database. The results are shown in Supplementary Figure 2C. Both gene sets have been refined by including more species and using a more stringent filtering approach, so the more recent database contains fewer and more conserved genes [4]. For the mollusca gene sets, a great improvement in completeness was observed between odb10 and odb12 (Supplementary Figure 2C); the metazoan completeness was marginally increased. Therefore, we evaluated all new assemblies produced since the first submission with the odb12 database.

(7) L107: How many scaffolds were obtained in total? After manual curation, how many of the scaffolds were placed in the "correct" chromosomes? How many scaffolds were in the shrapnel? Were these scaffolds mostly repetitive regions? Or did they contain important genetic information?

These are important questions. To evaluate the content of the “shrapnel”, we split the manually curated assembly into the 47 chromosomes and the 1840 residual scaffolds, and computed BUSCO scores for both. While the 47 chromosome scaffolds contain the majority of conserved genes: C:92.9%[S:92.7%,D:0.1%],F:4.0%,M:3.1% with metazoa_odb12 and C:88.7%[S:88.0%,D:0.7%],F:4.4%,M:6.9% with mollusca_odb12, the unplaced scaffolds still contain a few BUSCOs: C:2.5%[S:2.4%,D:0.1%],F:2.4%,M:95.1% from metazoa_odb12 and C:1.9%[S:1.7%,D:0.2%],F:1.2%,M:96.9% from mollusca_odb12. Even if only a few BUSCOs are present on these scaffolds, it means they contain important genetic information. Additionally, we observed low, but non-zero alignment of RNA reads to these scaffolds. We observed a slightly elevated repeat content in the unplaced scaffolds (Author response image 2), and a variable base composition (Figure 1C) compared to the chromosome scaffolds.

Quantification of repeat content in chromosome scaffolds and unplaced residual scaffolds.

Density plot showing fraction of repeat masked bases in total sequence length for chromosome scaffolds (i.e. scaffolds 1-47) in teal and all remaining small scaffolds (1840 scaffolds) in purple. Median repeat fraction is shown as vertical lines.

The slightly elevated repeat content in the unplaced scaffolds provides a likely explanation for their fragmented state: repeat-rich regions are inherently difficult to assemble and scaffold, as repetitive sequences cause ambiguous read alignments that prevent contigs from being confidently joined or anchored to chromosomal scaffolds during HiC-based scaffolding. This is consistent with the near-complete absence of BUSCO genes from the unplaced scaffolds - not because these fragments lack biologically relevant sequence entirely, as evidenced by the residual BUSCO hits and RNA read alignments, but because the gene-rich portions of the genome are largely captured in the 47 chromosome scaffolds. The unplaced scaffolds instead likely represent fragmented contigs from repetitive or low-complexity genomic regions, such as centromeres, telomeres, and transposable element clusters, where assembly graph complexity and collapsed repeats prevent confident placement. The variable base composition further supports this interpretation, as GC-extreme or low-complexity sequences are disproportionately represented in assembly shrapnel. Together, these observations suggest that the unplaced scaffolds contain limited unique coding content but reflect genuine repeat-rich genomic sequence that cannot currently be placed without additional long-range information, such as optical mapping or ultra-long reads.

(8) L33, 53, 240, 255, 279: Decapodiformes, not in italics.

We changed this throughout the text.

(9) L228: Can you put this expansion in perspective with other taxa?

We added a more detailed comparison of our gene family expansion with different species to the revised manuscript, as detailed in response 4 to reviewer 1.

(10) L251: "However, our results show how difficult it still is to assemble large genomes with high karyotype numbers." Can you clarify how your results show this, because it is equally spectacular to assemble the karyotype with only PacBio and Hi-C data (and no linkage mapping).

Indeed, it is correct that the recent improvements in data quality and scaffolding algorithms enable these “spectacular” chromosome-scale assemblies without the need for linkage mapping. This sentence reflected our expectation to resolve a clear karyotype as has been demonstrated for multiple cephalopod genomes in recent years, including two cuttlefish species (Octopus bimaculoides, Octopus vulgaris, Euprymna scolopes, Euprymna berryi, Acanthosepion lycidas and Acanthosepion esculenta). To our knowledge, none of these publications used linkage mapping or cytogenetic methods to confirm the karyotype. In this light, our resulting chromosome number and the discrepancy to a second assembly of the same species led us to this conclusion. We updated the section in the revised discussion as follows (lines 466-473):

“Taken together, our results illustrate the difficulty of assembling large genomes with high repeat content and large karyotypes, at least from sequencing data alone. Internal validation methods and genome comparisons across species are therefore important. Convergence of reliable estimates will, in turn, help identify chromosomal fusion-with-mixing events (FWM; fusion of two ancestral chromosomes followed by extensive shuffling of their gene content) that are clade specific. Early branching order in Decapodiformes has been notoriously unstable [53,84,94,144–147]; thus, such rare and irreversible FWM characters could be useful in further phylogenetic analysis of this clade [51,148].”

(11) L419: Why use the phased haplotype 1 instead of the primary assembly generated by hifiasm?

We thank the reviewer for this important question. We used the phased haplotype assembly because it provides a biologically coherent representation with the least amount of duplication by avoiding allele-collapsing and haplotype-switching that can be present in the primary assembly. We reasoned that this would result in clearer gene models and a more accurate representation of structural variation. However, we acknowledge that this comes at the cost of reduced contiguity and completeness, as becomes apparent in our BUSCO comparison shown in Supplementary Figure 2, where the phased haplotypes have fewer duplicated genes than the primary assembly, but more missing genes in turn. When reassembling both datasets for our comparison, we used the primary assembly to use the longest contigs as input for scaffolding.

(12) L444: It is unclear from what tissues and life stages RNA-seq data were used or were available from other species.

This is an important detail. RNA-seq data was collected from two adult Sepia officinalis, from various tissues (whole brain, retina, skin, mantle, arm, tentacle). For the long-read PacBio Isoseq data, tissue was taken from the animal used for genome sequencing (6 months old), and tissue for short-read Illumina RNA-seq was taken from another adult (8 months old). The data have been released on SRA (study accession SRP570862), where all sample details are listed as well. We added the SRA accession to the data availability section of the revised manuscript. We clarified the relevant sections in the methods:

lines 628-629:

“RNA was isolated from various flash-frozen tissues (different brain areas, mantle/epidermis, arm/tentacle; 5-10 mg each).”

lines 678-680:

“For short-read RNA sequencing, tissue from another animal (8-month-old adult, F0 from eggs collected in Normandie, France) was used. RNA was isolated from various flash-frozen tissues (different brain areas, skin and retina; 5 mg each).”

(13) L454, 469: Why is minimap2 in italics? It wasn't formatted like this before. Same for StringTie.

We thank the reviewer for their detailed methods review. In the updated methods section, all formatting of used softwares was harmonized.

(14) L461: Lophotrochozoa is a clade, not a genus or species. Not in italics.

This is now changed throughout the revised manuscript.

(15) Figure 1D: Axes labels are hard to read.

We have now increased the axis label size.

(16) Figure 2: Consider increasing font sizes. Many chromosome orientations seem to be flipped across species, which makes it harder to see smaller-scale rearrangements or notice less conserved chromosomes. Would it make sense to standardize these?

We increased the font sizes and plotted only fully collinear syntenic blocks (instead of aggregated syntenic regions, the default of GENESPACE) for improved readability.

References:

Below are references cited in our responses. References from the reproduced manuscript sections are included in the revised manuscript.

(1) Secomandi, S., Gallo, G.R., Rossi, R., Rodríguez Fernandes, C., Jarvis, E.D., Bonisoli-Alquati, A., Gianfranceschi, L., and Formenti, G. (2025). Pangenome graphs and their applications in biodiversity genomics. Nat. Genet. 57, 13–26. https://doi.org/10.1038/s41588-024-02029-6.

(2) Open2C, Abdennur, N., Fudenberg, G., Flyamer, I.M., Galitsyna, A.A., Goloborodko, A., Imakaev, M., and Venev, S.V. (2023). Pairtools: from sequencing data to chromosome contacts. Preprint at bioRxiv, https://doi.org/10.1101/2023.02.13.528389 https://doi.org/10.1101/2023.02.13.528389.

(3) Lupše, N., Reid, A., Taite, M., Kubodera, T., and Allcock, A.L. (2023). Cuttlefishes (Cephalopoda, Sepiidae): the bare bones—an hypothesis of relationships. Mar. Biol. 170, 93. https://doi.org/10.1007/s00227-023-04195-3.

(4) Tegenfeldt, F., Kuznetsov, D., Manni, M., Berkeley, M., Zdobnov, E.M., and Kriventseva, E.V. (2025). OrthoDB and BUSCO update: annotation of orthologs with wider sampling of genomes. Nucleic Acids Res. 53, D516–D522. https://doi.org/10.1093/nar/gkae987.

Version 2

Abstract

Coleoid cephalopods, a subclass of mollusks that includes octopuses, cuttlefish and squid, exhibit sophisticated biological features such as dynamic and neurally driven camouflage behavior, inter-individual communication, single-lens camera-like eyes, the largest brains among invertebrates and a distinctive embryonic development.

The common cuttlefish Sepia officinalis has served as a model organism in various research fields, spanning biophysics, neurobiology, behavior, evolution, ecology and biomechanics.

More recently, it has become a model to investigate the neural mechanisms underlying cephalopod camouflage, using quantitative behavioral approaches alongside molecular techniques to characterize the identity, evolution and development of neuronal cell types.

Despite significant interest in this animal, a high-quality, annotated genome of this species is still lacking. To address this, we sequenced and assembled a chromosome-scale genome for S. officinalis. Our assembly spans 5.68 billion base pairs and comprises 1n=47 repeat-rich chromosome scaffolds. This was unexpected because the haploid karyotypes of other decapods indicate 46 chromosomes. Detailed comparisons of our data to those from published decapod genome assemblies and to another recent genome assembly of S. officinalis (itself suggesting 1n=49 chromosomes) in fact revealed clear homologies between 46 scaffolds across all the datasets. The discrepancies between datasets are explained by highly repetitive regions, impairing proper read alignments. We conclude that the true karyotype of S. officinalis is probably 1n=46 chromosomes, a likely ancestral and if true, conserved decapod karyotype.

Our results include a comprehensive gene annotation and full-length transcript prediction, which we used to characterize orthologous gene families across mollusks. We identified several large-scale expansions specific to cephalopods, with many genes specific to neural or non-neural tissues of adult S. officinalis. In summary, this genome should provide a valuable resource for future research on the evolution, brain organization, information processing, development and behavior in this important clade.

Introduction

Coleoid cephalopods (octopus, squid, cuttlefish) are a highly derived group of mollusks, characterized by the largest nervous systems among all invertebrates (ca. 500 million neurons in adult octopus of which 200 million are in the central brain1,2, compared to ca. 140,000 in the fruit fly3 or 70 million in the mouse4) and specializations with a great historical importance for neuroscience (e.g., “giant axons”5 and “giant synapses”6–8). These animals exhibit very sophisticated behaviors such as dynamic camouflage9–15, learning16–18, social communication19–21 and hunting22–24 as well as two-stage sleep25–27. Because of these characteristics, cephalopods have been the focus of many fundamental studies by biologists, biophysicists and physiologists over the past century28–37. Thanks to advances in sequencing technologies, coleoid cephalopods are now also emerging as animal models in molecular neurobiology, evolution and genomics38. Recent studies examined cephalopod biology from the perspectives of single-cell gene expression39–42, genome topology and gene regulation43,44.

Despite recent technical progress in genomics, the genomes of cephalopods remain challenging to assemble because of their large sizes and high repeat fractions. While the genomes of Octopodiformes (Octopus, Eledone, Argonauta) are either smaller than (1.1 Gigabases or Gb45) or comparable in size to that of humans (around 3 Gb46,47), the typical genomes of Decapodiformes (squids and cuttlefish) often reach 6 Gb48,49. The biggest contributing factors to this genome expansion are transposable elements (TEs), with different TE classes differentiating squid and cuttlefish genomes from those of octopods50.

Besides such differences in genome sizes, karyotypes in coleoid cephalopods are also unusual when compared to those of other invertebrates, including other mollusks: for example, haploid chromosome numbers are around 30 in octopuses and around 46 in squids and cuttlefish49,51, while they are between 8 and 19 in non-cephalopod mollusks52–59. Karyotype reconstruction and validation techniques in animals with such large genomes are also not as developed as they are for more familiar invertebrate and vertebrate species.

The common cuttlefish Sepia officinalis has been used as a model organism for studies in neural control and development60,61, behavior9,17, evolution62, environmental studies63,64 and biomechanics65,66. Found in the Mediterranean, the English Channel and the European coastal Atlantic, it can produce large egg clutches containing hundreds of embryos, several times throughout its semelparous reproductive cycle67,68. More recently, this species has emerged as an important model to study the neural basis of cephalopod camouflage69,70 building on extensive behavioral studies9–15 and neuronal cell type characterization, evolution and development71–73.

Despite a widespread interest in this animal, including the recent sequencing of several cephalopod species as part of the Aquatic Symbiosis Genomics project74, a detailed, annotated genomic resource for this species is still lacking. We describe here the sequencing, assembly, and annotation of the Sepia officinalis genome.

Our data provided initial evidence for the existence of 1n=47 chromosomes. We compared them with available genome assemblies (i) from another S. officinalis individual by the Darwin Tree of Life project (DToL)75, itself suggesting 49 chromosomes, and (ii) from other decapod species such as Euprymna scolopes, Doryteuthis pealeii and Acanthosepion esculentum, each with estimates of 46 chromosomes49. When compared with the assemblies generated for the other species, both S. officinalis assemblies contained additional and independent chromosome splits. We thus investigated the repeat content and the alignment of raw data at the discrepant chromosome junctions to assess whether these differences were of a technical or biological nature.

Results

We sequenced the genome of a 6-month-old Sepia officinalis male individual, reared from eggs collected in the Portuguese Atlantic coast. The DNA was extracted from brain tissue and sequenced using long-read (PacBio HiFi) and chromatin conformation (Hi-C) methods (Figure 1A,B). The sex of the animal was confirmed by qPCR following a recently described protocol76.

Sepia officinalis assembly statistics and quality control.

A) Specimen of S. officinalis (credit: Stephan Junek, MPI for Brain Research). B) Overview of the genome assembly workflow. Genome size was estimated from short DNA reads (Illumina) using GenomeScope77,78. The primary assembly was generated from long DNA reads (PacBio Sequel II) and chromatin conformation capture (Hi-C) reads (Dovetail OmniC) with hifiasm167. Assembly was scaffolded with YAHS82 and residual small scaffolds were manually placed in chromosomes. C) Snail plot of chromosome-scale S. officinalis assembly generated using blobtools2191 showing scaffold statistics (e.g. number of scaffolds, median scaffold length N50), base composition and completeness measured using Benchmarking Universal Single-Copy Orthologs (BUSCO)80 against the metazoa_odb12 database. D) Hi-C heatmap showing the 47 chromosome-scale scaffolds with few sequences remaining in unplaced scaffolds. X and y-axes show the genome position in Mbp. The heatmap was generated using juicebox83, 0-7039 observed counts (balanced) are shown.

Genome size and heterozygosity

The haploid genome size of S. officinalis was estimated to be ∼5.14 Gb based on k-mer estimation from the short-read data, with a high repeat content of 54 %77,78. The heterozygosity rate was estimated to be 1.03 %, which is higher than in octopus genomes, yet moderate among marine invertebrates46,47,49.

The size of the scaffolded assembly was 5.68 Gb (Figure 1C), about 10% greater than our initial GenomeScope estimate. Indeed, the sizes of most published metazoan genome assemblies deviate by more than 10% from estimates, with proportionally greater deviations as assembly size increases79.

Completeness of the genome was assessed using Benchmarking Universal Single-Copy Orthologs (BUSCO80) with metazoa_odb12. It was 94.3% [single copy: 93.2%, duplicated: 1.2%], with 4.5% fragmented and 1.2% missing BUSCOs. The BUSCO score with mollusca_odb12 for the final assembly was 90.2% completeness [single copy: 89.1%, duplicated: 1.1%] with 4.9% fragmented and 4.9% missing BUSCOs.

Assembly-based karyotype

The assembly was generated from PacBio HiFi reads and chromosome conformation capture (Hi-C) reads with ca. 23-fold and 11-fold coverage, respectively. Scaffolding was performed using Hi-C reads, placing 95.8 % of bases in 47 scaffolds. The Hi-C heatmap in Figure 1D shows the chromatin contacts in clusters corresponding to these 47 scaffolds, suggesting that the 47 clusters each correspond to a single chromosome.

To further test the quality of our assembled S. officinalis karyotype, we used several scaffolding programs and manually curated the scaffolds using two independent approaches. The first rested on HapHiC81, a tool based on Hi-C data that uses allele information from primary assembly programs, and scaffolds assemblies with the constraint of an expected number of chromosomes. The resulting Hi-C contact maps with an input of 46, 47, 48, 49 and 50 (Supplementary Figure 1) supported 47 scaffolds: an input of 46 scaffolds resulted in one clear chromosome merger (blue arrowhead); higher input numbers resulted in false chromosome splits (black arrowheads, 48, 49 and 50 scaffolds).

Second, we manually curated the scaffolds (obtained with YAHS82) using JBAT83. These two approaches, conducted independently and both based on Hi-C signal, again converged on 47 chromosomal scaffolds.

Karyotype comparisons with other Decapods

Several chromosome-scale genome assemblies of decapod cephalopods have been published recently. From those studies, a haploid karyotype of 46 chromosomes seems to be common and conserved among Decapodiformes42,49,84 (Figure 3A). Thus, we sought to further investigate and confirm our estimated and different karyotype for S. officinalis.

Besides assembly-based karyotypes, several cephalopod species have been investigated using cytogenetic techniques. These studies, however, report widely varying estimates of chromosome numbers for decapods, reflecting the difficulty of resolving large (diploid) chromosome numbers in situ. For example, 1n=46 chromosomes has been reported for two species of cuttlefish (Acanthosepion esculentum and Acanthosepion lycidas) and three loliginid squids85; while 1n=34 chromosomes has been reported for Aurosepina arabica86 and 1n=24 chromosomes in Acanthosepion pharaonis87. In Sepia officinalis, a karyotype of 1n=52 has even been described from testis samples88.

Comparison with another chromosome-scale Sepia officinalis assembly

A chromosome-scale assembly for Sepia officinalis was released recently by the Wellcome Sanger Institute’s Darwin Tree of Life project75 (DToL, GCA_964300435.1). That genome was assembled from a male individual using high coverage PacBio Sequel II (∼51x) and Arima2 Hi-C (∼80x) data, with a final assembly size of 5.8 Gb. The haploid chromosome number was estimated to be 49. To compare both S. officinalis datasets directly, we downloaded the DToL data and created two new assemblies using the pipeline described above (hifiasm using PacBio HiFi and Hi-C data). The resulting assemblies were overall very similar, with the DToL assembly having a slightly higher contiguity (N50 length, see Table 1) and BUSCO completeness (Supplementary Figure 2A,B) due to their higher sequencing coverage.

After scaffolding with YAHS, both datasets reached the previously identified chromosome numbers (1n=47 for MPIBR and 1n=49 for DToL, Figure 2A,B). To further investigate this surprising discrepancy, we aligned both assemblies using Winnowmap89,90 to locate the differences between them (Figure 2C). We observed four “breakpoints” (BP) of chromosome scaffolds: one in the MPIBR assembly compared to DToL (BP1: DToL_5 = MPIBR_40+44) and three in the DToL assembly compared to MPIBR (BP2: DToL_31+40 = MPIBR_2, BP3: DToL_41+46 = MPIBR_6, BP4: DToL_44+45 = MPIBR_7). We also aligned the assemblies to the chromosome-scale genome of another cuttlefish Acanthosepion esculentum (1n=46, GCA_964036315.1). In this alignment, all four breakpoints were collinear with single A. esculentum chromosomes (Figure 2D).

Comparison of two Sepia officinalis chromosome-scale assemblies indicates chromosome number of 1n=46.

Datasets were collected from two S. officinalis animals, one as described in this study (MPIBR), the second by the Darwin Tree of Life consortium (DToL)75. Both datasets were assembled using a common pipeline (hifiasm and YAHS). A) Hi-C contact map of the MPIBR primary assembly, scaffolded using YAHS without manual curation. Assembled 47 chromosome scaffolds are shown as blue boxes. B) Hi-C contact map of DToL primary assembly, scaffolded using YAHS without manual curation, showing 49 assembled chromosome scaffolds as blue boxes. C) Whole-genome alignment of both scaffolded assemblies using Winnowmap289, showing DToL on x-axis and MPIBR on the y-axis. The 4 “breakpoints” of chromosomes in either of the assemblies (3 breaks in DToL chromosomes compared to MPIBR, 1 break in MPIBR compared to DToL) are highlighted in different colors. D) Ribbon diagram showing the four breakpoints from C) compared to the chromosome-scale assembly from another cuttlefish, Acanthosepion esculentum (1n=46). The color of breakpoints are the same in panels C+D.

Syntenic comparison of three decapod species.

A) Taxonomy of selected cephalopod species showing their genome size (in gigabases, Gb) and haploid chromosome numbers. Taxonomy information was downloaded from NCBI taxonomy browser, divergence times for Coleoidea and Decapodiformes from192 and for Sepiidae from94. B) Genome-wide syntenic relationship between chromosomes of E. scolopes49 (top), D. pealeii49 (middle) and S. officinalis (bottom). Colored braids connect syntenic regions across genomes, with chromosomes drawn to physical scale. Euprymna chromosomes 45 and 46 are not shown because they contain too few orthogroups. C) Detailed synteny of Sepia chromosomes 40 (magenta) and 43 (darkblue) shown, that are joined in the other species and cause the different haploid chromosome number in Sepia. Riparian plots were generated using GENESPACE v1.2.3175.

To better understand the potential cause of these divergent chromosome numbers, we analyzed the Hi-C and HiFi coverage in the breakpoint regions (Supplementary Figure 3A). First, we aligned the Hi-Fi reads to the scaffolds and extracted all alignments along the 200 kb terminal scaffold windows to find any notable drops in coverage, or reads spanning any of the scaffold junctions. We detected no spanning reads. This is not surprising given that no contigs were assembled at these sites, resulting in the observed scaffold junctions. More interestingly, we noted a ∼5-fold decrease in HiFi coverage along the DToL scaffold_40 (part of BP2) relative to its flanking regions, indicating a highly repetitive, low-mappability region at this boundary.

Next, we realigned the Hi-C data to the scaffolded assemblies using bwa-mem291 and extracted all trans HiC pairs (between-scaffold contacts) using pairtools92. We normalized trans HiC contacts to the scaffold length and compared contact rates between breakpoint scaffolds to the baseline contact rate (computed from pairs of scaffolds with a clear 1-to-1 match between assemblies), and the contact rate within scaffolds (intra-scaffold pairs) (Supplementary Figure 3B,C). The contact rates within breakpoints were consistently lower than within scaffolds, likely falling below the threshold to be merged during assembly.

However, the contact rates at three of four breakpoints (BP1, BP3, BP4) were significantly elevated above the genome-wide background distribution (empirical p = 0.010, 0.005, 0.005 respectively), suggesting that they may represent intra-chromosomal contacts disrupted by a misassembly. Notably, BP2 was not significant (empirical p = 0.170), likely due to the low coverage and mappability around the DToL scaffold_40 boundary. Considered jointly, the three DToL breakpoint scaffold pairs showed significantly higher trans contact rates than the background (Wilcoxon rank-sum, one-tailed, U = 1771, p = 0.004).

Lastly, we analyzed the repeat landscape around the 200 kb scaffold ends using RepeatMasker93 and the custom repeat library that we had generated for Sepia officinalis (described further below). Compared to control scaffolds of the same assembly, we observed consistently elevated repeat content at the breakpoint junctions (mean 71.5% vs 67.6% masked bases), with an enrichment of unclassified repeats (32.1% vs 30.0%), which could explain a repeat-driven assembly fragmentation or scaffolding failure. The BP2 DToL scaffold_40 junction window was 99.99% masked (99.2% unclassified repeats), providing a likely mechanistic explanation for both the HiFi coverage drop and the absence of a significant trans Hi-C signal at this breakpoint. Taken together, these analyses suggest that the different chromosome numbers across the two S. officinalis assemblies are due to technical reasons, caused by repeat-rich scaffold boundaries that impair HiFi and Hi-C read alignment and in turn, correct assembly in these regions.

Comparisons with other chromosome-scale Decapodiformes genomes

To further investigate why our initial estimate of Sepia officinalis’s chromosome numbers differred from those in other studies, we compared our assembly with the chromosome-scale genomes of two other decapod cephalopods, Euprymna scolopes43 (a sepiolid) and Doryteuthis pealeii49 (a loliginid), both described as having 46 chromosomes (Figure 3A).

Using orthogroups to perform linkage analysis, we could detect a clear chromosome homology between Doryteuthis and Sepia, and to a lesser extent between Euprymna and Sepia (Figure 3B). We observed some small-scale rearrangements between Euprymna and Doryteuthis, such as a fusion of chromosomes 24 and 40 from Doryteuthis in chromosome 2 of Euprymna, which has been also observed in other Sepiolids42,84. Dotplots showing detailed pairwise synteny comparisons are shown in Supplementary Figure 4 (Sepia to Doryteuthis) and Supplementary Figure 5 (Sepia to Euprymna). The inferred species tree as part of this analysis places Euprymna as sister to Sepia and Doryteuthis, matching recent phylogenetic analyses53,94 (note, however, that this tree was constructed without the inclusion of outgroup taxa, and therefore lacks a reliable root).

This comparison revealed that chromosomes 40 and 43 in Sepia are merged into one in Euprymna and in Dorytheutis (Figure 3C), as they were in the DToL Sepia officinalis assembly (Figure 2C). Together with the whole-genome alignments between the cuttlefish assemblies described earlier (Figure 2D), these results lead to the following parsimonious conclusion: that the karyotype for S. officinalis is 1n=46 chromosomes, as it is for other Decapodiformes.

We also performed a synteny comparison including the cuttlefish A. esculentum that was recently annotated by genome liftover from Acanthosepion pharaonis95. That assembly, constructed from a female individual, showed a low read coverage for a sex chromosome in a ZZ/Z0 sex determination system96. Our syntenic comparison indicated a strong homology between the inferred Z chromosome of A. esculentum and chromosome 46 of S. officinalis (Supplementary Figure 6A). As indicated above, we determined the sex of our S. officinalis specimen by replicating the analysis used to identify the Z chromosome in A. esculentum96. For this, we aligned short-reads (Illumina) from the same S. officinalis individual to the assembly and examined the normalized read coverage for each chromosome (Supplementary Figure 6B). In contrast to the low coverage observed in the female A. esculentum assembly (Supplementary Figure 6C), we observed no significant decrease in read coverage for chromosome 46, suggesting that our material came from a male animal. Additionally, we used a recently published genotyping protocol for cephalopods76 and performed qPCR on extracted genomic DNA from tissue samples, confirming the male sex of our sequenced individual.

Genome repeat landscape

After creating a custom repeat library for Sepia officinalis using RepeatModeler97, we masked the genome using RepeatMasker93, resulting in 71.17% masked bases. The categories of repeats are shown in Figure 4A. Most repeats were not characterized (39.65% of total bp) and presumably represent ancient repeats that diverged beyond recognition98.

Genome annotation for Sepia officinalis.

A) Annotation of repeat landscape of the S. officinalis genome, annotated using RepeatModeler97. Full repeat landscape is shown on the left, annotated repeats (excluding unclassified or simple repeats) are shown on the right. B-C) Quality control of gene annotation and comparison to two other cuttlefish species using OMArk117. Results shown for Acanthosepion lycidas (GCA_963932145.1, Ensembl Genebuild), Sepia officinalis (BRAKER, this study) and Acanthosepion pharaonis193 (BRAKER). Lophotrochozoa was used as the ancestral clade. B) Completeness assessed by the presence of genes conserved in the clade, classified as single or multiple copies (duplicated), or missing. C) Consistency assessed by the proportion of proteins placed in the correct lineage (consistent); placement in incorrect lineages randomly (inconsistent) or to specific species (contamination), or no placement in known gene families (unknown). D) Phylogenetic tree of 13 molluscan species used for analysis of gene families with Orthofinder122. Species are colored by clade: purple = coleoid cephalopods, blue = nautiloid (non-coleoid cephalopod), green = non-cephalopod mollusk. E) Heatmap of largest gene families (orthogroups from Orthofinder, with more than 100 genes in any species), ordered from largest gene count across all species on the left. Families with at least one gene in S. officinalis are depicted. Rows show gene counts for each species (color capped at 500 genes), columns show orthogroups and their annotation by eggNOG mapper120,121 or InterProScan119, if available. Clade colors match D).

Retroelements constituted the largest characterized repeat category (17.32%) followed by DNA transposons (5.92%); 5.83% were annotated as simple repeats. As observed in other Decapodiformes, the Sepia genome contained almost no short interspersed nuclear elements (SINEs), supporting the hypothesis that the SINE expansion observed in octopuses occurred independently in their lineage50.

Gene modeling and annotation

The genome was further annotated using BRAKER80,99–114 combining short-and long-read RNA-seq data and publicly available protein data from multiple molluscan species (Doryteuthis pealeii49, Euprymna scolopes115, Octopus bimaculoides49, Octopus vulgaris47, Nautilus pompilius116, and Pecten maximus54). A total of 18,663 gene models and 23,768 proteins were annotated.

The gene annotation was evaluated using BUSCO80 v5.5.0 in protein mode, showing high completeness of the annotation (metazoa_odb10: C:98.2%[S:77.3%,D:20.9%], F:1.0%, M:0.8%, n:954 and mollusca_odb10: C:81.5%[S:59.6%,D:21.9%], F:0.9%, M:17.6%, n:5295). We further checked the completeness and consistency of our gene models using OMArk117, and compared them to genome annotations for two other cuttlefish species, Acanthosepion pharaonis and Acanthosepion lycidas (Figure 4B). Completeness was assessed by comparing the annotated genes to conserved orthologs (present in 80% of extant species) in a given taxonomic clade (here the superorder Lophotrochozoa). The S. officinalis annotation missed 181 out of 2373 genes, which is higher than with A. lycidas (96 genes) but lower than with A. pharaonis (458 genes).

In a consistency assessment, where the presence of known gene families from the lineage is evaluated (Figure 4C), S. officinalis contains low proportions of taxonomically inconsistent or fragmented proteins (ca. 13%) similar to A. lycidas (9%), giving confidence in the annotation. In contrast, more than 50% of A. pharaonis proteins are labeled as inconsistent or fragmented, which could indicate annotation errors117.

Overall, the annotations contain different numbers of predicted proteins which may reflect differences in the annotation method and the reference data used. The genome of S. officinalis contains fewer proteins (23,768) than those of A. lycidas (35,949) or A. pharaonis (53,515). In comparison, two octopus genomes contain similar numbers of proteins (O. vulgaris: 30,134; O. bimaculoides: 29,037) and were produced using NCBI’s RefSeq118 pipeline, suggesting that the differences observed across cuttlefish proteomes are probably of technical instead of biological origin.

Lastly, we assigned orthology information to the S. officinalis proteome using InterProScan119 and eggNOG-mapper120 to aid the interpretability of the resource for future transcriptomic or proteomic studies. Overall, 89% of proteins (21,204 out of 23,768) received an annotation from InterProScan from at least one of their databases. 59% of proteins (14,126 out of 23,768) were annotated by eggNOG-mapper, reflecting the more stringent orthology filters and prioritization of full-length matches implemented in the program121.

Analysis of expanded gene families

We sought to investigate the S. officinalis gene annotation and place it in the context of gene repertoires from other cephalopod or molluscan species. First, we collected available genome annotations from 12 other molluscan species (Table 2) and clustered them using OrthoFinder v.3.1.0122, resulting in 23,658 orthogroups, hereafter named gene families.

First, we investigated 36 of the gene families that contain more than 100 genes in any of the species, with 17 of these families containing at least one gene of S. officinalis, that reflect large-scale gene family expansions (Figure 4E). We used the InterProScan and eggNOG-mapper annotations to infer functional roles of these genes, selecting the most common gene annotation as the name of the gene family.

The zinc finger C2H2-type transcription factors (TFs) were grouped into three of the large gene families, with the largest family (OG0000000) only present in decapod cephalopods. This likely reflects the largely independent expansions in the octopod and decapod lineages that date back to a burst of transposon activity ca. 25 million years ago46,48,49. The largest expansion across mollusks occurs in the cadherin-like family (OG0000001): 310 in S. officinalis, 283 in D. pealeii, 209 in A. lycidas, 102 in O. vulgaris, 55 in O. bimaculoides, with low but non-zero counts in bivalves (C. virginica, M. gigas). This profile is consistent with the protocadherin expansion first described in O. bimaculoides46 and subsequently shown to be present across cephalopods48,49,123.

HPGDS (OG0000005, hematopoietic prostaglandin D synthase) is a glutathione-S-transferase family member that catalyzes the conversion of prostaglandins, which have well-described roles in immune responses in vertebrates and insects124,125. This family shows a broad expansion in decapods, with a lesser expansion in octopods.

Additionally, members of the glutathione-S-transferase families have been co-opted as S-crystallins, structural proteins found in the lens of cephalopods that may, or may not, retain enzymatic functions126,127.

Two large families are mostly lineage-restricted. The RING-type zinc finger family (OG0000058) has 103 copies in S. officinalis and 26 in A. lycidas but is absent in all other species except for E. scolopes. Conversely, OG0000002 (unknown function) has 479 copies in E. scolopes and only a few copies in the other species. This interesting Sepiolid-specific expansion warrants further characterization.

We estimated gene family evolution rates using CAFE5128 for all families with less than 100 copies in any species (this excludes the families described above, as very large copy-number differences between species preclude likelihood calculations under the applied birth-death model). After comparing different model parameters, we chose a gamma model with three rate categories, allowing for evolutionary rate variation among gene families. Out of the 12,895 gene families analyzed, 1,813 showed a significant (p < 0.05) expansion or contraction in at least one of the species. We focused our analysis on the 30 most significantly expanded families; among them were several retrotransposon-associated domains that have expanded specifically in S. officinalis: five families carrying

Retrovirus-related Pol polyprotein domains, two Reverse transcriptase domain families, and four Ribonuclease H-like families (Supplementary Figure 7A). There was no coordinate-based overlap of the coding sequences with annotated TEs from the RepeatMasker output (Methods).

In addition to the three large gene families of C2H2 zinc finger expansions, 45 gene families containing this TF type showed a significant change in the CAFE5 analysis. Notably, eight of the significant gene families, as well as four of the largest gene families, were annotated as CCHC-type zinc fingers, which contain a “zinc knuckle” motif that is characteristic of retroviral nucleocapsid proteins129 and is functionally integrated in the genomes of several species, including humans130.

Some gene families without any relationship to retrotransposons were also expanded. For example, the UGT2A1-related family is a UDP-glucuronosyltransferase, a class of enzymes central to phase II detoxification and conjugation of metabolites, reported in other mollusks in the context of environmental chemical tolerance131, and in insects in the context of pigmentation132. We also detected a family of homeodomain-like proteins, representing an expansion of this important TF family.

Tissue-specific expression of expanded gene families

To place the identified gene families in a functional context, we profiled their expression in the bulk RNA-seq data (taken from multiple tissues of S. officinalis) used originally for gene modeling (Figure 5A). Principal component analysis (PCA) revealed the largest axis of variation in gene expression to separate brain tissues from peripheral tissues, with skin being the most transcriptomically distinct (Figure 5A), consistent with the high number of tissue-specific differentially expressed (DE) genes identified in non-neural tissues (Figure 5B). We identified the genes belonging to expanded families that were differentially expressed across tissues and enriched gene ontology133,134 (GO) terms for them to gain additional insight. The large families excluded from CAFE5 modelling and the significantly expanded families identified by CAFE5 were analyzed separately.

Expression of expanded gene families in tissue bulk RNA-seq data.

Bulk RNA-seq data collected from one adult S. officinalis from different brain tissues (optic lobes - yellow, basal lobes - turquoise, vertical and subvertical lobes - orange, posterior subesophageal mass - purple), retina (red) and skin (blue, from the dorsal mantle). Tissue color code is identical throughout the figure. A) Principal component analysis (PCA) of the data, showing the first 2 PCs, colored by tissue. B) Barplot showing number of differentially expressed (DE) genes (i.e. marker genes) for each tissue, calculated against all other tissues using DESeq2187. C) Largest gene families (orthogroups) with differential expression in bulk RNA-seq data. Dot size shows the number of DE genes for each tissue. Families with enriched GO terms are highlighted in grey. D)+E) Dotplots of enriched gene ontology133,134 (GO) terms for large gene families, enriched using clusterProfiler189 using a hypergeometric test. Dot size shows the number of expressed genes per family with this GO term, x-axis shows the percentage of expressed genes from all genes with this GO term. Dot color shows the adjusted p-value after Benjamini-Hochberg false discovery rate (FDR) correction. CC: cellular component, MF: molecular function, BP: biological process. F) Heatmap of z-scored expression of all DE genes from the largest gene families with enriched GO terms.

Eleven of the largest gene families were expressed in our data (Figure 5C) and five had enriched GO terms (Figure 5D,E). Among them, the cadherin family showed brain-restricted expression and GO terms related to cell–cell adhesion and calcium binding, consistent with their role in neuronal connectivity and circuit formation46,135. Two C2H2 zinc finger gene families were expressed in the optic and vertical/subvertical lobes of the brain and in the skin, with GO terms related to DNA-binding, transcriptional regulation or development. The RING-type zinc finger family was expressed specifically in the skin, with GO terms including zinc binding and ubiquitin protein ligase activity, the canonical function of RING-domain E3 ligases136. Genes of the HPGDS/S-crystallin family were expressed in the brain (basal and optic lobes and posterior subesophageal mass) and skin, with GO terms related to glutathione metabolism, matching their described enzymatic function. We did not find expression in the retina, which is expected given that S-crystallins are expressed in lentigenic cells of the eye42,137 and these cells were not included during sampling.

Among the 30 most significantly expanded families examined (out of 1,813 total), expression was widespread (20/30) and tissue-specific differential expression was common (17/30), suggesting that a substantial proportion of expanded paralogs represent functional coding sequences with specialized spatial deployment (Supplementary Figure 7B). Ten of the retrotransposon-associated families were differentially expressed in the brain (optic and vertical/subvertical lobes) and skin, arguing against these loci being inactive repeat fragments and supporting their inclusion as transcribed gene models. Two significantly expanded families showed both differential expression and enriched GO terms (Supplementary Figure 7C). The first was the UGT2A1-related family, which had the largest number of differentially expressed genes overall, with expression concentrated in the skin, retina and posterior subesophageal mass of the brain. Enriched GO terms matched the described enzymatic function for this family, namely UDP-glycosyltransferase activity. The second gene family was the homeodomain-like family with enrichment for DNA binding terms consistent with their role as transcription factors, and was preferentially expressed in the vertical and subvertical brain lobes with weaker expression in other areas.

Collectively, many differentially expressed genes from expanded families were restricted to specific tissues or brain subregions (Figure 5F and Supplementary Figure 7D), indicating that paralogs within an expanded family have adopted distinct spatial expression domains and possibly, specialized functions.

Discussion

This study presents a detailed, chromosome-scale genome assembly and annotation for the common cuttlefish Sepia officinalis, providing an additional important resource for comparative genomics, neurobiology, and evolutionary studies within cephalopods and mollusks. The assembly was derived from a combination of PacBio HiFi long reads, Dovetail Omni-C chromatin conformation capture, and multiple rounds of scaffolding and manual curation. The resulting assembly has a haploid genome size of 5.68 Gb, and contains 47 chromosome-like clusters. This number prompted us to perform detailed analyses of chromosome structure and homology across various decapod genomes, leading to a likely consensus of 46 chromosomes. This resource should be valuable for future efforts in cephalopod genomics, especially within the Decapodiformes, where large and repetitive genomes have hindered previous efforts.

Chromosome Number and Karyotypic Variation

Our assembly contains an estimated haploid chromosome number of 47, a result produced by two independent scaffolding approaches (YAHS82 and HapHiC81) and supported by Hi-C contact maps. This karyotype, however, differed from those reported for other decapod cephalopods, such as Euprymna scolopes and Doryteuthis pealeii49, each with 46 chromosomes. Syntenic comparisons across datasets revealed that chromosomes 40 and 43 in S. officinalis correspond to a single chromosome in these other species, suggesting that the apparent split in Sepia may result from a technical artefact.

We also compared our results to the recently published S. officinalis genome from the Darwin Tree of Life project (DToL)75, which proposed 49 chromosomes for that species. This difference was surprising: two genome assemblies from the same species are usually expected to have the same karyotype. The discrepancy is attributable to one chromosome split in our assembly and three splits in the DToL assembly, and could stem from technical differences of the Hi-C methods (Dovetail Omni-C vs. Arima v2), sequencing depth (11-fold vs. 83-fold coverage) or the tissue used (optic lobe vs. eye). We investigated these splits, or breakpoints in two ways: first, we reassembled both datasets with the same parameters and aligned both to the genome of another cuttlefish A. esculentum (1n=46). In this alignment, all four breakpoints were collinear with single A. esculentum chromosomes, providing a phylogenetically grounded prediction of 1n=46 for Sepia officinalis. Second, we investigated read depth at the breakpoint junctions, and saw a significant increase of Hi-C read pairs spanning the junctions compared to random chromosome pairings. Together with a higher repeat content at the breakpoints, these analyses suggest a technical cause for the differences between assemblies, caused by repeat-rich scaffold boundaries that impaired HiFi and Hi-C read alignment and in turn, made correct assembly challenging in these regions.

In conclusion, the two independent genome assemblies and our analysis suggest three possible karyotypes for Sepia officinalis: 46, 47, or 49. The most likely explanation based on chromosome synteny is that the true number is 46, matching the numbers established for other cuttlefish and decapod species42,49,84. 47 chromosomes (our initial estimate) and 49 chromosomes (DToL estimate) would therefore be erroneous values explained by technical factors related to repeat content and low alignment rates preventing complete assembly.

Incorporating ultra-long read data may help to correctly assemble these problematic regions, as is now common for telomere-to-telomere assemblies138.

Another, but less likely, explanation for the observed differences could be variations in chromosomal architecture of different Sepia officinalis populations, due to idiosyncratic individual chromosome fusion. Intraspecific chromosome number variation is rare but not unprecedented; for instance, chromosomal polymorphism has been described in the butterfly Leptidea sinapis across different populations139,140. Notably, the two S. officinalis individuals used for sequencing originated from different regions - the Portuguese Atlantic (this study) and the French Mediterranean (DToL project), raising the possibility of geographic variation in chromosome structure. Future population-level analyses will hopefully determine whether S. officinalis exhibits karyotypic polymorphism.

Additional methods such as cytogenetic karyotyping or optical mapping such as BioNano141 (imaging of fluorescently tagged, linearized DNA) could be used to validate chromosome numbers. However, whereas karyotypes of octopuses have been consistent throughout the literature (1n=30)142,143, those measured in decapods vary greatly. For example, 1n=46 chromosomes have been reported for two species of cuttlefish (A. esculentum and A. lycidas) and three loliginid squids85; 1n=36 has been reported for A. arabica86 and 1n=24 in A. pharaonis87. In S. officinalis, a karyotype of 1n=52 is reported for testis samples88. Combining cytogenetic preparations with fluorescent labeling of centromeric or telomeric sequences, as demonstrated in the octopus A. aerolatus143 could help resolve these issues. Establishing a routine staining protocol would enable comprehensive tests at the species-and population-level.

Taken together, our results illustrate the difficulty of assembling large genomes with high repeat content and large karyotypes, at least from sequencing data alone. Internal validation methods and genome comparisons across species are therefore important. Convergence of reliable estimates will, in turn, help identify chromosomal fusion-with-mixing events (FWM; fusion of two ancestral chromosomes followed by extensive shuffling of their gene content) that are clade specific. Early branching order in Decapodiformes has been notoriously unstable53,84,94,144–147; thus, such rare and irreversible FWM characters could be useful in further phylogenetic analysis of this clade51,148.

In addition to studying chromosomal topology in phylogenetic reconstructions, some of the most interesting aspects of these rearrangements relate to changes of and innovation in regulatory elements that underlie phenotypic diversity. In coleoid cephalopods, it is thought that an ancient large-scale genome rearrangement was combined with lineage-specific changes and repeat expansions48–50. This restructuring gave rise to hundreds of tightly linked, evolutionarily unique microsyntenies, corresponding to distinct topological compartments with specialized regulatory architectures that contribute to complex, tissue-specific expression patterns in the nervous system and elsewhere43. Extending this, chromosomal conformation analyses in E. scolopes revealed that co-regulated eye and light-organ genes cluster at topologically associating domain (TAD) boundaries, and that an evolutionarily recent rearrangement at the dachshund (DAC) locus may have been instrumental in the emergence of the symbiotic light organ in Euprymna - directly linking specific chromosomal topology to morphological innovation44.

To understand the broader functional impact of these changes across coleoids, a recent study investigating Micro-C, RNA-seq, and ATAC-seq data from multiple species revealed broadly conserved chromatin domains, but also many lineage-specific chromatin loops that form novel regulatory signatures and impact expression profiles across species and tissues149.

Despite the observed small-scale regulatory changes, the chromosomes of decapods are considered to be more closely related to the ancestral coleoid karyotype than those of octopods. The derived octopod karyotype becomes apparent when comparing it to the genome of the vampire squid, an early-branching octopodiform (sister to all octopods) which retained features of the decapod, ancestral karyotype150. Taken together, the conserved karyotype of decapods accommodates fine-scale regulatory diversity that might underlie morphological diversity among species, which suggests that many regulatory innovations are still being evolutionarily explored through rearrangements within the existing chromosomes.

Genome Size and Repeat Landscape

The size of the genome of S. officinalis is estimated to be 5.68 Gb - comparable to those of other Decapodiformes and roughly twice the size of typical Octopodiformes genomes. We found that over 71% of the genome is repetitive, the repeats being dominated by unclassified elements and retrotransposons. The near absence of SINEs reinforces the idea that their expansion in octopuses was a lineage-specific event50. These results underscore the evolutionary complexity and dynamism of cephalopod genomes, shaped by waves of transposon activity that likely played a role in the evolution of novel traits43,49.

Gene Annotation and Comparative Assessment

We annotated 18,663 gene models and 23,768 proteins using a combination of short-and long-read RNA-seq data and molluscan protein references. Compared with other annotated cuttlefish genomes (A. lycidas, A. pharaonis), our annotation is conservative concerning gene numbers but stands out in terms of completeness and taxonomic consistency.

OMArk-based evaluations confirmed that our gene models have low levels of fragmentation and contamination, and high lineage-specific coherence. Striking differences across species in the number of predicted genes, duplications, and missing orthologs likely reflect variation in the annotation pipelines and reference data rather than true biological differences. Note that the comprehensive annotation of protein isoforms remains challenging even for model organisms151,152. Still, our dataset provides a solid foundation that may be refined with additional (long-read) transcriptomic data taken from diverse tissues and developmental stages.

We characterized the S. officinalis gene models further by clustering them into families together with 12 other cephalopod and non-cephalopod mollusks. As in other coleoid genomes, we observed large-scale expansions of gene families such as C2H2 zinc finger transcription factors and protocadherins, genes implicated in regulatory innovation, neural development and plasticity46,48,49,123. These conserved expansions may be linked to the unusual cognitive complexity and behavioral sophistication of these invertebrates.

We profiled the expression of expanded gene families in bulk RNA-seq data from multiple S. officinalis tissues and found that many genes were differentially expressed, frequently in a tissue-restricted manner. This spatial partitioning of paralogs across tissues validates many expanded families as likely functional and biologically relevant, and suggests that subfunctionalization or neofunctionalization may have accompanied gene family expansion in S. officinalis, with individual family members acquiring specialized roles153,154. Note that we only investigated the expression of the largest copy number families, and the most significantly expanded families. Whether this pattern generalizes to the broader set of expanded families remains to be determined.

Among the large families, the cadherin and C2H2 zinc finger families showed preferential expression in neural tissues, consistent with known roles in synaptic organization and transcriptional regulation of neural development, respectively. The skin-specific expression of the RING-type zinc finger family is notable given its restricted expansion in decapods.

RING domains are characteristic of E3 ubiquitin ligases that catalyze protein ubiquitination and target substrates for proteasomal degradation155,156. Their expression could hint at unique demands in protein turnover in the skin, but the specific function of this gene family in cuttlefish skin remains to be determined.

Many of the smaller but significantly expanded gene families, which were predominantly found in decapod genomes, were linked to retrotransposons, like retrovirus-related Pol polyprotein, reverse transcriptase domain, and ribonuclease H-like families. We confirmed that these genes were genuine coding sequences and not artifacts of TE repeats, and that they were expressed in S. officinalis tissues. The combination of coding-sequence annotation, expression, and tissue specificity is consistent with at least some of these loci having been retained as functional genes after retroelement-related origins.

Notably, we observed many expanded families containing CCHC-type zinc fingers, which contain a “zinc knuckle” motif, characteristic of retroviral nucleocapsid proteins129. In addition to the retroviral function, some eukaryotic proteins containing CCHC-domains have been described, which likely originate from domesticated nucleocapsid sequences and play important roles in RNA metabolism130,157. Taken together, these gene families may represent a recent wave of retroelement activity, and the variation in copy number across different Decapodiformes is consistent with repeated recruitment and/or independent retention of retroelement-derived sequences in different decapod lineages.

Two non-retroviral gene families were also differentially expressed in S. officinalis tissues: a UDP-glucuronosyltransferase (UGT) family and a homeodomain-like family. The expression of the UGT family, particularly in skin and retina is interesting: enzymes of the UGT family are involved in pigment metabolism in insects132, but have also been reported in mollusks as playing roles in environmental chemical tolerance131. The homeodomain-like family contains transcription factors with important roles in body patterning and brain regionalization158. We found members of this family to be expressed in the vertical and subvertical lobes of the brain, but also more weakly in other brain areas, suggesting novel regulatory roles of this transcription factor class in higher-order neural circuits in brain areas associated with learning and memory in cephalopods18.

A potential caveat of this analysis is that the gene modeling approaches used for the species were different: while the majority of genomes were annotated with the RefSeq pipeline118, others used Ensembl genebuild159, BRAKER80,99–114 or other custom approaches. This may have introduced artificially inflated gene counts due to insufficient isoform resolution, or missing gene models due to the absence of reference data. In the future, as genome annotations become progressively refined (as for D. pealeii160), gene family-level analyses will become more powerful and enable the identification of additional lineage-or species-specific innovations.

Conclusions and Outlook

This study provides a new genomic resource for Sepia officinalis. The chromosome-scale assembly and annotation open the door to in-depth studies of gene regulation, neural circuit development, and genome evolution across coleoid cephalopods.

As we move toward a more complete picture of cephalopod genome evolution, the integration of chromosomal synteny, regulatory architecture, and transcriptomic diversity across species will be especially important. The S. officinalis genome represents an important step on this path, enabling high-resolution comparative and functional analyses of one of the most enigmatic and evolutionarily successful invertebrate lineages.

Materials & Methods

Animal husbandry

All research and animal care procedures were carried out following the institutional guidelines that comply with national and international laws and policies (DIRECTIVE 2010/63/EU; German animal welfare act; FELASA guidelines). European cuttlefish Sepia officinalis were hatched from eggs collected in the Portuguese Atlantic and reared in a seawater system, at 20°C. The closed system contains 4,000 L of artificial seawater (ASW; Instant Ocean) with a salinity of 33‰ and pH of 8–8.5. Water quality was tested daily and adjusted as required. Trace elements and amino acids were supplied weekly. Marine LED lights above each tank provided a 12/12-h light/dark cycle with gradual on-and off-sets at 07:00 and 19:00. The animals were fed live food (either Hemimysis spp. or small Palaemonetes spp.) ad libitum twice per day. The animals were housed together in 120 L glass tanks with a constant water through-flow resulting in five complete water exchanges per hour. Enrichment consisted of natural fine-grained sand substrate, seaweed (Caulerpa prolifera), rocks of different sizes and various natural and man-made three-dimensional objects.

Tissue preparation

Animals underwent terminal anesthesia to prevent animal suffering or distress, following the Guidelines for the Care and Welfare of Cephalopods in Research161,162. Animals were transferred to a bucket containing ethanol 2% (v/v) in ASW. After 5 minutes, animals were gently probed for simple avoidance reflexes, and the ethanol concentration in the ASW was gradually increased to a maximum of 5%. Sufficient depth of anesthesia was reached when the animal no longer reacted even to stronger stimuli (touching and pinching with tweezers), and no reaction was observed when a hand was moved in the visual area and the cornea was touched163. In this deep anesthesia, the animals were decapitated and the tissue was rapidly transferred into ice-cold, calcium-free Ringer solution (460 mM NaCl, 10 mM KCl, 51 mM MgCl2, 10 mM glucose, 2 mM Glutamine, 10 mM HEPES, pH 7.4) bubbled with oxygen. Brain and body tissue was dissected under a stereoscope and flash-frozen in liquid nitrogen before storage at-80°C until further use.

Unless stated otherwise, sequencing libraries were prepared from the same individual (6-month-old adult Sepia officinalis, F1 from eggs collected in Portugal).

Long-read Whole Genome Library Preparation and Sequencing

The long-read library was prepared and sequenced at the MPI for Plant Research in Cologne, Germany. Genomic DNA was extracted from flash-frozen brain tissue with the MagAttract HMW DNA kit (Qiagen, Cat. no. 67563) and the sequencing library was prepared with the SMRTbell express template prep kit 2.0 (PacBio, Cat. no. 100-938-900). The library was sequenced on 5 SMRT cells on the PacBio Sequel II with the Sequel II binding kit 2.2 (102-089-000), Sequel II sequencing kit 2.0 (101-820-200) and SMRT Cell 8M tray (101-389-001).

Long-read RNA library preparation and sequencing

RNA was isolated from various flash-frozen tissues (different brain areas, mantle/epidermis, arm/tentacle; 5-10 mg each). First, tissue samples were homogenized in TRIzol using a homogenizer (VDI12/S12N-5S, VWR, Germany). RNA was isolated and DNAseI treated according to the Direct-zol RNA Mini prep kit from ZymoResearch (R2050). RNA was quantified by spectrophotometry (Nanodrop, Thermo Scientific, USA) and its quality was assessed with a RNA 6000 Nanochip on Bioanalyzer (Agilent Technologies, Germany).

The Iso-Seq libraries were prepared and sequenced at the Sequencing facility of the MPI for Plant Research, using a method targeting the 5’ cap and poly-A tail of mRNAs, described in detail in164. Briefly, mRNA was pooled from the 8 tissue samples and cDNA was synthesized using the TeloPrime Full-Length cDNA Amplification Kit V2 (Lexogen, Cat Nr: 013.08 or 013.24). The sequencing library was prepared with the SMRTbell express template prep kit 2.0 (PacBio, Cat. no. 100-938-900). The pooled library was sequenced on 1 SMRT cell on the PacBio Sequel II system using the Sequel II binding kit 2.1 (101-843-000), Sequel II sequencing kit 2.0 (101-820-200) and SMRT Cell 8M tray (101-389-001).

Further, a separate library was prepared from optic lobe cDNA and sequenced on a second SMRT cell using the same reagents described above.

Omni-C Library Preparation and Sequencing

An Omni-C library was prepared and sequenced at the MPI for Plant Research in Cologne, Germany. The library was prepared from brain tissue using the Dovetail Omni-C™ Kit (Dovetail Genomics, Cat. No. 21005) according to the manufacturer’s protocol. Briefly, chromatin was fixed in place in the nucleus. Fixed chromatin was digested with DNaseI then extracted. Chromatin ends were repaired and ligated to a biotinylated bridge adapter followed by proximity ligation of adapter-containing ends. After proximity ligation, crosslinks were reversed and the DNA was purified from proteins. Purified DNA was treated to remove biotin that was not internal to ligated fragments. The sequencing library was generated using Illumina-compatible adapters. Biotin-containing fragments were isolated using streptavidin beads before PCR enrichment of the library. The library was sequenced on an Illumina NextSeq 2000 platform to generate 400 million 2 x 150 bp read pairs.

Short-read DNA sequencing

High molecular weight genomic DNA was isolated from brain tissue (optic lobe) using NEB Monarch gDNA Purification Kit (T3010S). First, 10 mg of flash-frozen tissue were submerged in 100 µl of tissue lysis buffer, cut into small pieces and transferred into a 1.5 ml tube containing an additional 100 µl of lysis buffer. Then, 3 µl of 20 U/µl Proteinase K was added to the tissue sample and incubated for 2 h at 56°C in a thermal mixer with agitation until tissue pieces were completely dissolved. After incubation, tissue lysate was purified from the debris by centrifugation for 3 min at maximal speed (Eppendorf, Germany). Clean supernatant was transferred to a new 1.5 ml tube and incubated with 3 µl RNase A for 5 min at 56°C with agitation at full speed. After RNase A incubation tissue lysate was column purified and eluted with 80 µl of elution buffer according to the manufacturer protocol. The purity of gDNA was assessed using A260/280 and A260/230 absorbance ratios measured by NanoDrop spectrophotometer (ThermoFisher Scientific). The integrity of the obtained gDNA was checked using 0,75% agarose gel electrophoresis by loading 100 ng of gDNA sample and using HindIII digested Lambda DNA (NEB#3012) as a marker.

DNA sequencing libraries were prepared from 500 ng of high MW gDNA using Illumina DNA PCR-Free Tagmentation Library Prep Kit (20041794) and UD Indexes Set A (20026121).

Obtained dual-indexed single-stranded libraries were quantified using the Qubit single-stranded DNA (ssDNA) assay kit (Q10212, ThermoFisher Scientific). 1000 pM of the final library were run on P3 reagents and Illumina NextSeq2000 using Illumina DNA PCR-Free Sequencing and Indexing primer (20041797) with 151 bp paired-end reads.

Short-read RNA library preparation and sequencing

For short-read RNA sequencing, tissue from another animal (8-month-old adult, F0 from eggs collected in Normandie, France) was used. RNA was isolated from various flash-frozen tissues (different brain areas, skin and retina; 5 mg each). First, tissue samples were homogenized in TRIzol using a homogenizer (VDI12/S12N-5S, VWR, Germany) following DNAseI treatment and column purification according to the Direct-zol RNA Micro prep kit from ZymoResearch (R2062). The RNA integrity and quantity were measured with the Qubit fluorometer (Invitrogen, Q33216) and the 2100 Bioanalyzer (Agilent Technologies, Germany).

RNA-seq libraries were prepared from 300 ng of total RNA, using the Illumina TruSeq stranded mRNA library prep kit (Cat:20020594) and IDT for Illumina xGen UDI-UMI Adapters (Cat:10005903). Libraries were sequenced on an Illumina NextSeq500 (Mid output, 300 cyc) and NextSeq2000 (P3, 300 cyc) with 145 bp paired-end reads.

Nuclear genome assembly

Before assembly, PacBio HiFi reads were subjected to additional trimming to remove residual adapter sequences using NCBI’s VecScreen tool. The k-mer distribution was estimated using Meryl165 within the Merfin166 package with a k-mer size of 21, and genome size was estimated using GenomeScope77 from Illumina short reads and PacBio HiFi data. The primary assembly was generated with hifiasm167 using a combination of HiFi and Hi-C reads. Scaffolding was performed iteratively with YAHS82 on the phased haplotype 1. We adjusted the run parameters for scaffolding resolutions (-r 1000,2000,5000,10000,20000,30000,50000,70000,100000,150000,200000,250000,300000, 350000,400000,500000,700000,1000000,2000000,5000000,10000000,15000000,20000000,30000000,40000000,50000000,60000000,70000000,80000000,100000000,110000000,120 000000,150000000,170000000,200000000,500000000), repetitions per resolution (-R3), minimum read mapping quality (-q1) and telomeric sequence (--telo-motif “TTAGGG”) from the only experimentially determined telomere motif for cephalopods (Amphioctopus areolatus143, accessed via TeloBase168) to optimize the tool’s performance for a large, fragmented assembly. Finally, the assembly was manually curated using JBAT83 to place residual scaffolds into chromosome scaffolds. Different versions of the assembly were evaluated based on Hi-C coverage, mRNA alignment and completeness based on BUSCO80 v5.5.0 (metazoa_odb10 and mollusca_odb10).

Mitochondrial genome assembly

To assemble the mitochondrial genome, we aligned a published S. officinalis mitochondrial genome169 (NC_007895.1) to the PacBio Hi-Fi and Omni-C reads using minimap2170. The hits were extracted using seqtk171 subseq and assembled using hifiasm167. The resulting contigs contained multiple repeats of the circular mitochondrial genome, and were aligned back to NC_007895.1 to extract the final sequence.

Nuclear genome annotation

Repetitive elements in the genome were softmasked using RepeatMasker v4.1.7-p193 (with rmblast v2.14.1+) (-xsmall and-gff options) and after creating a custom repeat library with RepeatModeler v.2.0.697 (without LTRstruct option).

Gene models were created using BRAKER380,99–114, run via the Docker container, on the softmasked genome using both RNA-seq and protein data. We used long-read Iso-Seq data and short-read Illumina RNA-seq data from various tissues (brain, skin, mantle, retina) generated for this study (see above). For protein data, publicly available proteomes for Doryteuthis pealeii49, Euprymna scolopes115, Octopus bimaculoides49, Octopus vulgaris47, Nautilus pompilius116, Pecten maximus54 were used for training, as well as a previously generated proteome for Sepia officinalis from StringTie gene models (this study, described below). RNA-seq data was input into BRAKER3 using the --bam option, protein files were specified with the --prot_seq option. Untranslated regions (UTRs) were added by the-addUTR=on parameter. The BUSCO completeness of the resulting gene set was maximized using TSEBRA within BRAKER on the BUSCO lineage metazoa_odb10.

For the initial protein set used as input for BRAKER, the short and long RNA reads were aligned to the genome using minimap2170. Transcript models were predicted with StringTie v3.0.0172 using the --conservative and --mix options. The resulting GTF files were combined using the transcript merge mode, resulting in a set of non-redundant transcripts. Finally, TransDecoder v5.7.0173 was run with default parameters to translate coding regions in the transcripts.

The final gene annotations for S. officinalis were assessed for completeness using OMArk v.0.3.0117 with the ancestral clade Lophotrochozoa and without including splice information, accessed via their webserver. Proteins were annotated with orthology information using InterProScan119 v5.73-104 including lookup of annotations and GO terms (options-iprlookup-goterms). Further orthology information was added using the eggNOG-mapper v2.1.12120 webserver with the eggNOG v5.0 database and default parameters.

Whole genome alignment and synteny analysis

For whole genome alignments, the assembly produced for Sepia officinalis in this study and the published assembly from the Darwin Tree of Life project (GCA_964300435.1) or the Acanthosepion esculentum genome (GCA_964036315.1) were aligned using Winnowmap289 and visualized with a custom script in R v4.4.2174.

For synteny analyses, proteome and gtf files were downloaded for Doryteuthis pealeii and Euprymna scolopes. Annotation files for Acanthosepion esculentum were recently generated by liftover annotation from Acanthosepion pharaonis96 and downloaded from Zenodo95.

Synteny analyses between all chromosomes of the compared species were performed using the R package GENESPACE v.1.2.3175 with default parameters, described briefly below.

Protein sequence similarity was first estimated using DIAMOND2109 in fast mode, and orthogroups and pairwise orthologues were inferred using OrthoFinder v2.5176 with hierarchical orthogroups (HOGs) enabled. Prior to synteny inference, tandem arrays were condensed to their most central representative gene, and gene rank order was recalculated on these array-representative genes to reduce confounding effects of tandem duplication on collinearity detection.

Syntenic blocks were identified pairwise between all genome combinations using MCScanX177, constrained to DIAMOND hits where both query and target genes belonged to the same orthogroup (onlyOgAnchors = TRUE). Initial anchor hits were clustered into large syntenic regions using a density-based spatial clustering approach (dbscan178), with a minimum block size of five anchor genes (blkSize = 5) and a maximum of five intervening non-anchor genes permitted within a block (nGaps = 5). Anchor clustering used a search radius of 25 gene-rank positions (blkRadius = 25). All hits falling within a syntenic buffer of 100 gene-rank positions around confirmed block anchors (synBuff = 100) were retained as syntenic. No secondary syntenic hits were included (nSecondaryHits = 0). Syntenic orthogroups were integrated across all pairwise comparisons and collapsed into a pan-genome annotation anchored to S. officinalis as the reference genome.

Syntenic relationships were visualized as riparian plots and pairwise dotplots using the built-in plotting functions of GENESPACE v1.2.3. Riparian plots were constructed using physical chromosomal coordinates (useOrder = FALSE) with S. officinalis as the reference, displaying all three genomes. A second riparian plot was generated highlighting a region of interest. Pairwise dotplots were produced for the S. officinalis–D. pealeii and S. officinalis–E. scolopes genome comparisons, displaying only synteny-validated hits (type = “syntenic”) with a minimum synteny score of 10 (minScore = 10) and a minimum of 10 genes per chromosome pair required for display (minGenes2plot = 10).

Coverage analysis

For the analysis of breakpoints between S. officinalis assemblies, Hi-C data was aligned using bwa-mem2 v2.391 and quantified using pairtools v1.1.092. Contacts between scaffolds (trans pairs) were extracted from deduplicated read pairs (pair type UU). For each of the four breakpoints, all trans pairs between the two flanking scaffold halves were extracted using awk. As controls, the corresponding joined scaffold from the opposing assembly was paired with the nearest-length uninvolved scaffold from the same assembly, and trans pairs were extracted identically.

Genome-wide background and intra-scaffold contact distributions were computed in a single pass over each pairs file using awk, recording the total number of trans pairs and intra-scaffold pairs (>1 kb separation) for all scaffold pairs across the genome. Trans contact rates were normalized by the product of the two scaffold lengths (pairs per Mb²) to correct for length bias. For visualization, genome-wide trans pair positions were binned into 500 kb windows along each scaffold to produce contact density tracks. To test whether trans contact rates at breakpoints were elevated above background, empirical p-values were computed as the fraction of all genome-wide scaffold pair rates equal to or exceeding the observed breakpoint rate. For the three DToL breakpoints jointly, a one-tailed Wilcoxon rank-sum test was applied against the background distribution.

HiFi reads were aligned to the scaffolded assemblies using minimap2170 and duplicate-marked alignments were removed. HiFi read depth was computed over the terminal 200 kb at each scaffold end using pysam v0.22.1179 count_coverage with a mapping quality threshold of ≥10, binned at 1 kb resolution. Spanning reads - defined as reads with supplementary alignments crossing a scaffold boundary - were identified by querying split alignments at each junction using pysam.

Repeat content at scaffold junctions was characterized by extracting 200 kb windows flanking each breakpoint and the corresponding correct scaffold termini, and running RepeatMasker v4.1.7-p193 (with rmblast v2.14.1+) against a de novo repeat library generated by RepeatModeler v.2.0.697 from the same assembly (described above). Repeat annotations were parsed and collapsed into eight classes (SINE, LINE, DNA transposons, LTR elements, simple repeats, low-complexity regions, unknown repeats, and other), and the masked fraction per class was quantified for each window.

For sex chromosome analysis, read coverage across chromosomes was analyzed as described recently96. Short reads were aligned using STAR v2.7.11b180 to our chromosome-scale assembly. For A. esculentum, short reads (ERR12945500) and assembly (GCA_964036315.1) were downloaded from NCBI. The sequencing depth was calculated using mosdepth181 by a window size of 500,000 bp and normalized to the median coverage of the first chromosome. Chromosomes with significantly reduced read coverage were identified by a one-sided Wilcoxon rank-sum test of each chromosome’s normalized depth windows against all remaining chromosomes. P-values were corrected using the Benjamini-Hochberg method and considered only for chromosomes with at least 10% decrease in median normalized depth.

Gene family expansion analysis

Orthogroups were inferred across 13 molluscan species (Table 2), including S. officinalis, using OrthoFinder v3.1.0122 with default parameters. The input proteomes included the longest protein isoform per gene for each species. The rooted species tree from OrthoFinder182,184 was converted to an ultrametric tree using the R package ape183 v5.8.1.

Gene families were filtered by removing orthogroups present in only a single species, and by separating orthogroups containing 100 or more gene copies in any species, as extreme copy-number differences in gene families prevent likelihood calculation under the applied birth-death model.

Gene family evolution rates were estimated using CAFE5128 v5.1.1 on the filtered orthogroups, using the ultrametric species tree as input. Four models were evaluated: the base model (single global lambda), and Gamma models with k = 2, 3, and 4 rate categories, which allow evolutionary rate variation among gene families. The Gamma k = 3 model was selected based on the best (lowest) final log-likelihood score. All subsequent statistical inferences were performed under this model.

For families showing statistically significant expansion or contraction (p < 0.05 after Bonferroni correction), branch-specific copy-number changes were extracted from the CAFE5 output. Families were categorized as S. officinalis-specific, coleoid-specific, or broad expansions based on the distribution of significant changes across the phylogeny.

To assess whether expanded gene families in S. officinalis contained genes derived from or embedded within repetitive elements, a coordinate-based overlap analysis was performed. For each gene in an expanded orthogroup, the overlap between its coding sequence (CDS) coordinates and RepeatMasker annotations was computed using bedtools intersect v2.30185. To avoid double-counting when multiple repeat annotations overlapped the same coding bases, overlapping repeat intervals were merged per gene prior to summing covered bases, and the overlap fraction was computed as merged covered bases divided by total CDS length.

Bulk RNA-seq analysis

Quality-filtered paired-end RNA-seq reads were aligned to the S. officinalis genome assembly using STAR180 (v2.7.11b) with the following parameters: --outSAMmultNmax 1 (retaining only uniquely mapping reads), --outFilterIntronMotifs RemoveNoncanonical (removing reads with non-canonical splice junctions), and --outSAMtype BAM SortedByCoordinate.

Gene-level read counts were quantified using featureCounts186 from the Subread package (v2.0.8) with the gene annotation described above (GTF format). Counting was performed at the exon level (-t exon) and summarized by gene (-g gene_id), using paired-end mode (-p--countReadPairs) to count fragments rather than individual reads. Only reads with mapping quality ≥ 255 (-Q 255), corresponding to uniquely mapped reads in the STAR output, were included in the count matrix. The resulting gene count matrix was used as input for differential expression (DE) analysis.

Tissue-specific marker genes were identified using DESeq2187 (v1.42.0) with a one-vs-all comparison strategy. For each tissue, samples were grouped into a binary condition (“target” vs. “other”), where “target” represented the tissue of interest and “other” comprised all remaining tissues. Raw count matrices were filtered to retain genes with ≥10 counts in at least 3 samples prior to analysis. DE testing was performed using the Wald test, log2 fold changes were shrunk using the apeglm188 method. Genes were classified as tissue markers if they met the following criteria: adjusted p-value (Benjamini-Hochberg FDR) < 0.05, log2 fold change > 1, and mean expression in the target tissue > 5.

To generate gene-level GO133,134 annotations for the S. officinalis genome, GO term assignments from the InterProScan output (detailed earlier) were used. First, transcript IDs were collapsed to gene IDs and for genes with multiple transcripts, GO terms were aggregated using a union strategy, collecting all unique GO terms across all isoforms and annotation sources to maximize coverage. GO terms present in fewer than 5 genes or more than 500 genes in the expressed gene universe were excluded from enrichment testing to reduce noise from overly specific or generic terms.

GO enrichment analysis was performed using the clusterProfiler189,190 (v4.12.6) enricher function with custom GO annotations generated as described above. For each gene set (DE genes for each tissue), enrichment was tested against the background universe of all expressed genes (genes passing the low-count filter in DESeq2). Over-representation was assessed using the hypergeometric test, with p-values adjusted for multiple testing using the Benjamini-Hochberg procedure. GO terms with adjusted p-value (FDR) < 0.05 were considered significantly enriched.

S. officinalis members of each expanded gene family were cross-referenced against tissue-specific marker genes identified by DESeq2 (as described earlier). For families with at least one differentially expressed member, GO enrichment was performed using enricher() from clusterProfiler as described for tissue marker genes. For families with at least one enriched GO term, expression patterns of differentially expressed members were visualized as a heatmap of z-scored VST values, with samples grouped by tissue.